\documentclass{article}

% --- Conference style (ICLR 2027 target) ---
% Using NeurIPS 2026 template as placeholder until ICLR 2027 template is released
% Will convert to ICLR template in September 2026
\usepackage[nonatbib]{neurips_2026}

% --- Encoding & typography ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}

% --- Math ---
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{nicefrac}

% --- Tables ---
\usepackage{booktabs}
\usepackage{multirow}

% --- Figures ---
\usepackage{graphicx}
\usepackage{subcaption}

% --- References & links ---
\usepackage{hyperref}
\usepackage{url}
\usepackage[capitalize,noabbrev]{cleveref}
\usepackage[round]{natbib}

% --- Other ---
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{thmtools}

% --- Theorem environments (shared counter) ---
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% --- Complexity classes ---
\newcommand{\TC}{\mathrm{TC}^0}
\newcommand{\AC}{\mathrm{AC}^0}
\newcommand{\NC}{\mathrm{NC}^1}
\newcommand{\PTIME}{\mathrm{P}}
\newcommand{\NP}{\mathrm{NP}}
\newcommand{\coNP}{\mathrm{coNP}}
\newcommand{\PSPACE}{\mathrm{PSPACE}}
\newcommand{\BPP}{\mathrm{BPP}}
\newcommand{\IP}{\mathrm{IP}}
\newcommand{\AM}{\mathrm{AM}}

% --- Paper-specific macros ---
\newcommand{\VC}{\mathrm{VC}}           % Verification Complexity
\newcommand{\GC}{\mathrm{GC}}           % Generation Complexity
\newcommand{\GVGap}{\Delta_{\mathrm{GV}}} % Generation-Verification Gap
\newcommand{\calM}{\mathcal{M}}         % Model class
\newcommand{\calF}{\mathcal{F}}         % Task family
\newcommand{\calV}{\mathcal{V}}         % Verifier
\newcommand{\calD}{\mathcal{D}}         % Distribution
\newcommand{\calC}{\mathcal{C}}         % Complexity class
\newcommand{\CoT}{\mathrm{CoT}}
\newcommand{\poly}{\mathrm{poly}}
\newcommand{\negl}{\mathrm{negl}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\Pr}{\mathbf{Pr}}

\title{On the Verification Complexity of LLM Outputs:\\ When Checking is as Hard as Generating}

\author{Anonymous}
\date{}

\begin{document}
\maketitle

% ============================================================
\begin{abstract}
% ============================================================
The assumption that verifying a language model's output is easier than generating it underpins reward modeling, self-consistency decoding, and scalable oversight proposals.
We show this assumption is not uniformly valid by developing a \emph{verification complexity framework} that classifies reasoning tasks by the computational complexity of checking candidate solutions.
We define verification complexity classes for LLM tasks, formalize the generation-verification gap, and prove three main results:
(1)~a \emph{verification advantage theorem} showing that inference-time mitigations succeed if and only if the task's verification complexity falls within the augmented model's capability class;
(2)~a \emph{self-consistency condition} characterizing exactly when majority voting improves accuracy, connecting the Condorcet jury theorem to complexity-theoretic verification;
and (3)~a \emph{gap collapse theorem} demonstrating that for hierarchical planning tasks, verification is provably as hard as generation under standard complexity assumptions, rendering reward-model-based training ineffective.
We map verification complexity onto six reasoning gap types, producing testable predictions for GSM8K, MATH, PlanBench, and 3-SAT benchmarks.
The framework provides a principled answer to when AI oversight is tractable and when it is not.
\end{abstract}

% ============================================================
\section{Introduction}
\label{sec:intro}
% ============================================================

A pervasive assumption in modern AI development is that \emph{verification is easier than generation}.
This belief manifests across the stack: reward models learn to distinguish good from bad outputs~\citep{christiano2017deep,ouyang2022training}; self-consistency methods assume correct answers converge under repeated sampling~\citep{wang2023selfconsistency}; and scalable oversight proposals rely on humans or weaker models checking the work of stronger systems~\citep{burns2023weak,irving2018debate}.

The assumption has a venerable pedigree in computational complexity theory.
The class $\NP$ is precisely the set of decision problems for which candidate solutions can be \emph{verified} in polynomial time, even when \emph{finding} solutions may require exponential time --- the conjectured $\PTIME \neq \NP$ separation.
Yet $\NP$ is not all of computation.
For problems in $\PSPACE$ (e.g., planning under uncertainty), verification of optimality can be as hard as generation.
For problems requiring interactive proofs, the verification power depends on the protocol structure: $\IP = \PSPACE$ with interaction~\citep{shamir1992ip}, but only $\NP$ without.

Despite this rich classical theory, \textbf{no prior work formally connects verification complexity to the landscape of LLM reasoning and training methods}.
Song et al.~(\citeyear{song2024mind}) define a generation-verification gap operationally (as the accuracy difference between generation and verification on held-out problems) but do not ground it in complexity theory.
Stechly et al.~(\citeyear{stechly2024self}) show empirically that LLMs cannot self-verify planning solutions, without explaining \emph{why} from a complexity-theoretic perspective.
Brown-Cohen et al.~(\citeyear{browncohen2024debate}) connect debate protocols to $\PSPACE$ verification, but only for alignment --- not for routine output checking.

We bridge this gap with three contributions:
\begin{enumerate}[leftmargin=*,itemsep=2pt]
    \item A \textbf{verification complexity framework} that classifies LLM reasoning tasks by the computational complexity of their verification procedures, connecting to classical complexity classes (\Cref{sec:framework}).
    Theorem~1 (Verification Advantage) formalizes a foundational observation --- that verification-based re-ranking succeeds iff the verification problem is within the augmented model's capability class --- which, while implicit in the literature, has not been stated or proved as a formal complexity-theoretic result.
    Theorem~3 (Gap Collapse) synthesizes known results on HTN planning complexity~\citep{erol1996complexity,behnke2024verification} into a single statement with direct implications for LLM oversight; the novelty lies in the connection to the verification framework, not the underlying complexity results.
    \item \textbf{Theorem~2 (Self-Consistency Condition)}, our main technical contribution, which introduces the \emph{computational bottleneck} formalism (\Cref{def:bottleneck}) and uses it to characterize when majority voting improves accuracy.
    The key result is part~(c): the non-monotone relationship between verification complexity and error correlation, which predicts that tasks with \emph{easier} verification can exhibit \emph{higher} error correlation than tasks with harder verification, depending on whether the bottleneck is shared or stochastic.
    \item A \textbf{verification complexity table} mapping six reasoning gap types to their verification properties, producing testable predictions (\Cref{sec:taxonomy,sec:experiments}).
\end{enumerate}

The practical implications are immediate: for tasks with polynomial-time verification (mathematical reasoning, code with test suites, constraint satisfaction), current verification-based methods are well-founded.
For tasks where verification is hard (hierarchical planning, optimization under uncertainty, open-ended reasoning), these methods are provably limited, and no amount of scale will overcome the barrier.

% ============================================================
\section{Background}
\label{sec:background}
% ============================================================

\paragraph{Transformer expressiveness.}
Fixed-depth log-precision transformers compute within $\TC$~\citep{merrill2022saturated}, a subclass of $\NC$.
With polynomial-length chain-of-thought (CoT), expressiveness extends to $\PTIME$~\citep{merrill2024expressive,li2024chain}.
This hierarchy defines the \emph{generation capability} of different model configurations.

\paragraph{Verification in complexity theory.}
$\NP$ is defined by polynomial-time verifiability: $L \in \NP$ iff there exists a polynomial-time verifier $V$ and polynomial $p$ such that $x \in L \Leftrightarrow \exists w, |w| \leq p(|x|), V(x,w) = 1$.
Interactive proof systems~\citep{goldwasser1985knowledge,babai1985trading} generalize this: a polynomial-time probabilistic verifier interacts with an unbounded prover over multiple rounds.
Shamir~(\citeyear{shamir1992ip}) proved $\IP = \PSPACE$, showing that interaction dramatically extends verification power.
The PCP theorem~\citep{arora1998proof} shows $\NP$ proofs can be checked by reading only $O(1)$ bits.

\paragraph{LLM verification methods.}
Self-consistency~\citep{wang2023selfconsistency} samples multiple reasoning paths and takes the majority answer, implicitly assuming a Condorcet-like independence condition.
Process reward models~\citep{lightman2023lets} verify individual reasoning steps.
Generative verifiers~\citep{hosseini2024genrm} reframe verification as generation, producing chain-of-thought critiques.
All assume that some form of tractable verification exists for the target task.

% ============================================================
\section{Verification Complexity Framework}
\label{sec:framework}
% ============================================================

We develop a formal framework connecting the verification complexity of reasoning tasks to the effectiveness of LLM training and inference methods.

\begin{definition}[Reasoning Task]
\label{def:task}
A \emph{reasoning task} is a tuple $\calF = (X, Y, V, D)$ where $X$ is an input space, $Y$ is an output space, $V : X \times Y \to \{0, 1\}$ is a decidable verification function, and $D$ is a distribution over $X$.
For input $x \in X$, a solution $y \in Y$ is \emph{correct} if $V(x, y) = 1$.
\end{definition}

The verification function $V$ is the central object. Its computational complexity determines what verification-based methods can achieve.

\begin{definition}[Verification Complexity Class]
\label{def:vc}
For a reasoning task $\calF = (X, Y, V, D)$, the \emph{verification complexity} $\VC(\calF)$ is the complexity class of the language $L_V = \{(x, y) : V(x, y) = 1\}$.
We say $\calF$ is \emph{efficiently verifiable} if $\VC(\calF) \subseteq \PTIME$.
\end{definition}

\begin{example}
\label{ex:sat}
For SAT, a candidate assignment $y$ can be verified in $O(n \cdot m)$ time by evaluating each clause, so $\VC(\text{SAT}) = \PTIME$.
For MAX-SAT (``does $y$ satisfy at least $k$ clauses and no assignment satisfies $k{+}1$?''), verifying optimality requires proving no better assignment exists, so $\VC(\text{MAX-SAT}_{\text{opt}}) = \coNP$.
\end{example}

\begin{definition}[Generation Complexity]
\label{def:gc}
The \emph{generation complexity} $\GC(\calF)$ is the complexity class of the search problem: given $x \sim D$, find $y$ such that $V(x, y) = 1$ (or determine that no such $y$ exists).
\end{definition}

\begin{definition}[Generation-Verification Gap]
\label{def:gvgap}
The \emph{generation-verification gap} of task $\calF$ is the pair $\GVGap(\calF) = (\GC(\calF), \VC(\calF))$.
We say $\calF$ has a \emph{positive gap} if $\VC(\calF) \subsetneq \GC(\calF)$ under standard complexity assumptions, and a \emph{collapsed gap} if $\VC(\calF) = \GC(\calF)$.
\end{definition}

The classical $\PTIME \neq \NP$ conjecture asserts that $\NP$-complete problems have a positive gap: verification is in $\PTIME$ while generation is $\NP$-hard.
But not all reasoning tasks are in $\NP$ --- and not all verification is in $\PTIME$.

\begin{definition}[Model Capability Class]
\label{def:cap}
A model class $\calM$ has \emph{capability class} $\mathrm{cap}(\calM) = \calC$ if $\calM$ can compute exactly the functions in $\calC$.
For fixed-depth transformers, $\mathrm{cap}(\calM_{\text{ff}}) = \TC$.
For transformers with unbounded CoT, $\mathrm{cap}(\calM_{\CoT}) = \PTIME$.
For transformers with tool access (e.g., code execution), $\mathrm{cap}(\calM_{\text{tool}})$ extends beyond $\PTIME$ depending on the tool.
\end{definition}

\begin{definition}[Effective Verification]
\label{def:effver}
A model class $\calM$ can \emph{effectively verify} task $\calF$ if $\VC(\calF) \subseteq \mathrm{cap}(\calM)$.
That is, the model has sufficient computational power to run the verification procedure.
\end{definition}

This definition captures the key insight: a reward model, a self-consistency checker, or a human overseer can only reliably verify outputs when the verification problem itself falls within their computational capabilities.
A $\TC$-bounded transformer cannot verify solutions to problems whose verification requires $\PTIME$ computation --- even if the problem has a positive generation-verification gap in the classical sense.

\begin{definition}[Computational Bottleneck]
\label{def:bottleneck}
Let $\calF = (X, Y, V, D)$ be a reasoning task, $\calM$ a model class, and $V(x,y) = g(V_1(x,y), \ldots, V_k(x,y))$ a decomposition of the verification function into subtasks for some Boolean combination $g$.
A subtask $V_j$ is a \emph{computational bottleneck} for $(\calF, \calM)$ if it satisfies:
\begin{enumerate}[label=(B\arabic*)]
    \item \textbf{Necessity.} There exists $\alpha > 0$ such that
    \[
    \Pr_{x \sim D}\!\bigl[\,\exists\, y_1, y_2 \in Y :\ V(x,y_1) \neq V(x,y_2)\ \land\ V_i(x,y_1) = V_i(x,y_2)\ \forall\, i \neq j\,\bigr] \geq \alpha.
    \]
    That is, on at least an $\alpha$-fraction of inputs, the remaining subtasks $\{V_i\}_{i \neq j}$ do \emph{not} suffice to determine $V(x,\cdot)$: there exist two candidate answers that $V$ distinguishes but every subtask other than $V_j$ fails to distinguish.
    This means $V_j$ is genuinely \emph{required}, not merely informative.
    \item \textbf{Hardness.} $V_j \notin \mathrm{cap}(\calM)$: the model class $\calM$ cannot compute $V_j$.
\end{enumerate}
We write $V_{\mathrm{sub}} \coloneqq V_j$ for the bottleneck subtask.
A bottleneck $V_{\mathrm{sub}}$ is \emph{shared} if $V_{\mathrm{sub}}$ is required on all inputs where the model errs --- formally, if for every $x$ in the support of $D$ with $\Pr[V(x, \calM(x)) = 0] > 0$, distinguishing $\calM(x)$ from a correct answer requires evaluating $V_{\mathrm{sub}}$.
A bottleneck is \emph{stochastic} if the set of inputs requiring $V_{\mathrm{sub}}$ has measure $q \in (0,1)$ under $D$, and whether a particular input $x$ requires $V_{\mathrm{sub}}$ is determined by instance-specific structure (e.g., random combinatorial properties) rather than a limitation common to all inputs.
\end{definition}

\begin{example}
For algorithmic tasks (Type~4, e.g., longest increasing subsequence), the bottleneck $V_{\mathrm{sub}}$ is the algorithm itself (e.g., dynamic programming for LIS). Every instance requires this algorithm, so the bottleneck is \emph{shared}: all models lacking the algorithm fail on the same instances, producing high error correlation.
For 3-SAT near the phase transition (Type~5), the bottleneck is the combinatorial search required to verify unsatisfiability of hard instances.
Whether a particular random formula is hard depends on its clause structure, so the bottleneck is \emph{stochastic}: different models (or sampling runs) may succeed or fail on different instances.
\end{example}

\begin{remark}[Non-uniqueness of the Decomposition]
\label{rem:decomposition}
The decomposition $V(x,y) = g(V_1(x,y), \ldots, V_k(x,y))$ in \Cref{def:bottleneck} is not unique: the same verification function $V$ may admit multiple valid decompositions into subtasks.
Our results hold for \emph{any} decomposition satisfying the stated conditions.
If a computational bottleneck exists under one valid decomposition, the correlation bound of \Cref{lem:verification_correlation} follows.
In practice, the ``natural'' decomposition corresponds to the algorithmic steps required for verification (e.g., for HTN planning: check action sequence feasibility, then check decomposition-tree existence).
A canonical notion could be defined via circuit complexity --- taking the minimal decomposition (fewest subtasks, or minimal total circuit size) --- but this would introduce additional complexity-theoretic overhead without strengthening the main results.
We leave the study of canonical decompositions to future work.
\end{remark}

% ============================================================
\section{Main Results}
\label{sec:theorems}
% ============================================================

\subsection{The Verification Advantage Theorem}

Our first result formalizes when verification-based inference methods are effective.

\begin{theorem}[Verification Advantage]
\label{thm:advantage}
Let $\calF$ be a reasoning task and $\calM$ a model class augmented with strategy $S$ (e.g., CoT, tool use, reward model re-ranking).
Denote the augmented capability class $\mathrm{cap}(\calM + S)$.
Then:
\begin{enumerate}[label=(\alph*)]
    \item If $\VC(\calF) \subseteq \mathrm{cap}(\calM + S)$ and the augmentation $S$ includes access to a correct verifier, then for any $\epsilon > 0$, there exists $N = \poly(|x|, 1/\epsilon)$ such that sampling $N$ candidates and selecting the one that passes verification achieves accuracy $\geq 1 - \epsilon$, provided the generator produces a correct answer with probability $p > 0$.
    \item If $\VC(\calF) \not\subseteq \mathrm{cap}(\calM + S)$, then no polynomial-time verification-based re-ranking strategy can improve accuracy beyond $p + \negl(|x|)$, where $p$ is the generator's base accuracy.
\end{enumerate}
\end{theorem}

\begin{proof}
\textbf{Part (a).}
Since $\VC(\calF) \subseteq \mathrm{cap}(\calM + S)$, the augmented model can compute $V(x, y)$ for any candidate $(x, y)$.
Sample $N$ candidates $y_1, \ldots, y_N$ from the generator independently.
The probability that all $N$ candidates are incorrect is $(1-p)^N$.
Setting $(1-p)^N \leq \epsilon$ and solving gives $N \geq \lceil \log(\epsilon) / \log(1-p) \rceil = O(\log(1/\epsilon) / p)$, which is polynomial in $|x|$ and $1/\epsilon$ when $p \geq 1/\poly(|x|)$.
Selecting any $y_i$ with $V(x, y_i) = 1$ achieves the claimed accuracy.

\textbf{Part (b).}
Suppose for contradiction that a polynomial-time verification-based re-ranking strategy $S'$ achieves accuracy $p + \delta$ for some non-negligible $\delta = 1/\poly(|x|)$ on distribution $D$, where $S'$ operates within $\mathrm{cap}(\calM + S)$.
Then $S'$ implicitly distinguishes correct from incorrect candidates: given input $x \sim D$ and a candidate $y$, run $S'$ to re-rank candidates; $S'$ selects a correct $y$ with probability at least $p + \delta$.

We use this to construct a randomized algorithm $A$ for $L_V = \{(x,y) : V(x,y) = 1\}$.
On input $(x,y)$, algorithm $A$ generates $N-1$ additional candidates from the generator and adds $y$ to the pool, then runs $S'$ to select the ``best'' candidate.
If $S'$ selects $y$, output $V(x,y) = 1$; otherwise output $0$.
Since $S'$ preferentially selects correct candidates (achieving accuracy $p + \delta > p$), this procedure has advantage $\delta$ over random guessing at deciding membership in $L_V$.

By standard amplification (taking the majority vote of $O(1/\delta^2)$ independent runs of $A$), we obtain a $\BPP$ algorithm for $L_V$ with error probability $< 1/3$.
By Adleman's theorem ($\BPP \subseteq \PTIME/\poly$), this yields a polynomial-size nonuniform algorithm for $L_V$.
For verification problems arising from $\NP$-hard or $\coNP$-hard tasks, $L_V \notin \PTIME/\poly$ under standard assumptions (e.g., the polynomial hierarchy does not collapse).
This contradicts $\VC(\calF) \not\subseteq \mathrm{cap}(\calM + S)$.
The $\negl(|x|)$ term accounts for the probability of guessing correctly.
\end{proof}

\begin{remark}
\Cref{thm:advantage} explains why best-of-$N$ re-ranking with a reward model succeeds on mathematical reasoning (where $\VC \subseteq \PTIME$) but fails on open-ended planning.
It also explains why process reward models~\citep{lightman2023lets} are effective: step-level verification decomposes a $\PTIME$-verification problem into $O(n)$ subproblems, each individually easier to learn.
\end{remark}

\subsection{The Self-Consistency Condition}

Self-consistency~\citep{wang2023selfconsistency} samples $N$ reasoning paths and takes the majority answer.
We characterize exactly when this works.

\begin{theorem}[Self-Consistency Condition]
\label{thm:selfconsistency}
Let $\calF$ be a reasoning task with unique correct answer $y^*$ for each input $x$.
Let $\calM$ be a model that, on input $x$, produces answer $y_i$ with probability $p_i$ independently across $N$ samples, where $p^* = p_{y^*}$ is the probability of the correct answer.
Then majority voting over $N$ samples satisfies:
\begin{enumerate}[label=(\alph*)]
    \item \textbf{(Convergence)} If $p^* > \max_{y \neq y^*} p_y$ (the correct answer is the plurality), then majority accuracy $\to 1$ as $N \to \infty$, with convergence rate $\exp(-\Omega(N))$.
    \item \textbf{(Independence requirement)} If samples share a systematic bias --- i.e., errors are positively correlated with correlation $\rho > 0$ --- then the effective sample size is $N_{\text{eff}} = N / (1 + (N-1)\rho)$, and convergence slows to $\exp(-\Omega(N_{\text{eff}}))$.
    \item \textbf{(Bottleneck structure)} Let $V_{\mathrm{sub}}$ be a computational bottleneck (\Cref{def:bottleneck}) for $(\calF, \calM)$.
    \begin{enumerate}[label=(\roman*)]
        \item \textbf{(Shared bottleneck)} If $V_{\mathrm{sub}}$ is shared, occurring with probability $q > 0$ under $D$, and the conditional error probability satisfies $r = \Pr[\calM(x) \text{ incorrect} \mid V_{\mathrm{sub}} \text{ required at } x] > 1/2$, let $p_0 < r$ be the base error rate absent the bottleneck and $p_e = rq + p_0(1-q)$ the marginal error rate.
        Then:
        \[
        \rho \;\geq\; q(1-q)(r - p_0)^2 \,/\, [p_e(1-p_e)] \;>\; 0.
        \]
        Self-consistency provides limited benefit: $N_{\mathrm{eff}} = O(1/\rho)$ regardless of~$N$.
        \item \textbf{(Stochastic bottleneck --- qualitative prediction, not formally proved)} If $V_{\mathrm{sub}}$ is stochastic --- i.e., whether input $x$ requires $V_{\mathrm{sub}}$ depends on instance-specific combinatorial structure --- then we conjecture that \emph{between-model} error correlation satisfies $\rho_{\mathrm{between}} = O(q^2)$, where $q$ is the bottleneck probability, since different models with different learned heuristics face different hard instances.
        This is consistent with the empirical finding of $\rho = 0.06$ for 3-SAT near the phase transition (\Cref{sec:discussion}).
        Self-consistency across diverse models can improve accuracy.
        \item \textbf{(Within-model dominance)} For within-model sampling at fixed temperature from a single model $M$, the shared-bottleneck regime dominates: all $N$ samples use the same weights and share the same computational limitations, so $\rho_{\mathrm{within}} > 0$ and $N_{\mathrm{eff}}$ is bounded independently of $N$.
    \end{enumerate}
\end{enumerate}
\end{theorem}

\begin{proof}
\textbf{Part (a)} follows from the strong law of large numbers applied to the indicator $\mathbf{1}[y_i = y^*]$.
Since $p^*$ is the unique maximum, for $N$ sufficiently large, $y^*$ is the plurality with probability $1 - \exp(-\Omega(N))$ by Hoeffding's inequality.

\textbf{Part (b).}
Under the equicorrelated model where $\mathrm{Corr}(\mathbf{1}[y_i = y^*], \mathbf{1}[y_j = y^*]) = \rho$ for $i \neq j$, the variance of the sample mean is $\mathrm{Var}(\bar{Y}) = p^*(1-p^*)(1 + (N-1)\rho) / N$.
This is equivalent to $N_{\text{eff}} = N / (1 + (N-1)\rho)$ independent samples.
When $\rho = \Omega(1)$, $N_{\text{eff}} = O(1/\rho)$ regardless of $N$, and majority voting converges to a fixed (potentially incorrect) answer.

\textbf{Part (c).}
The error correlation $\rho$ depends on the \emph{structure} of the computational bottleneck (\Cref{def:bottleneck}), not solely on $\VC(\calF)$.
We prove each sub-part.

\emph{Part (c.i): Shared bottleneck.}
If $\VC(\calF) \not\subseteq \mathrm{cap}(\calM)$, then by \Cref{lem:vc_to_bottleneck} (Appendix~\ref{app:proofs}), there exists a computational bottleneck $V_{\mathrm{sub}}$ with occurrence probability $q > 0$ and conditional error rate $r > 1/2$.
\Cref{lem:verification_correlation} then yields $\rho \geq q(1-q)(r - p_0)^2 / [p_e(1 - p_e)] > 0$.
Since all $N$ within-model samples share the same weights, they face the same bottleneck on the same inputs, so the shared-bottleneck bound applies: $N_{\mathrm{eff}} = N/(1 + (N-1)\rho) = O(1/\rho)$.

\emph{Part (c.ii): Stochastic bottleneck (qualitative argument).}
Part~(c.ii) is an empirically-supported prediction rather than a formally proved bound.
When the set of inputs requiring $V_{\mathrm{sub}}$ depends on instance-specific combinatorial structure (e.g., random clause structure in 3-SAT near the phase transition), different models with different learned heuristics succeed or fail on different instances.
Between-model error indicators $\mathbf{1}[M_a(x) \text{ incorrect}]$ and $\mathbf{1}[M_b(x) \text{ incorrect}]$ are conditionally independent given $x$ (since the models are independent), and the between-instance correlation is low because which instances are hard varies stochastically.
Heuristically, if the overlap between hard instances for $M_a$ and $M_b$ is $O(q^2)$ (the probability that a random instance is independently hard for both), then $\rho_{\mathrm{between}} = O(q^2)$.
A formal derivation would require a model of how different architectures and training runs produce different bottleneck instance sets, which we leave to future work.
The prediction is consistent with the empirical finding of $\rho = 0.06$ for 3-SAT near the phase transition (\Cref{sec:discussion}).

\emph{Part (c.iii): Within-model dominance.}
For within-model sampling from a single model $M$ at fixed temperature $T > 0$, all $N$ samples share the weights of $M$.
If $V_{\mathrm{sub}} \notin \mathrm{cap}(M)$ (condition B2 of \Cref{def:bottleneck}), then on every input $x$ where $V_{\mathrm{sub}}$ is required, \emph{every} sample from $M$ faces the identical computational limitation.
Sampling randomness (temperature) perturbs token-level choices but cannot bypass a missing computational capability.
Hence within-model errors satisfy the shared-bottleneck regime of part~(c.i), and $N_{\mathrm{eff}}$ is bounded independently of $N$.
\end{proof}

\begin{corollary}
\label{cor:selfconsistency_prediction}
Within-model self-consistency yields diminishing returns on tasks where the model faces a shared structural bottleneck (algorithmic gap, depth limitation), regardless of whether $\VC$ is tractable.
Between-model ensembles are most effective on tasks with stochastic, instance-specific difficulty (e.g., 3-SAT near the phase transition), where different models fail on different instances.
\end{corollary}

\begin{remark}[Graceful Degradation]
\label{rem:degradation}
The correlation bound in \Cref{thm:selfconsistency}(c.i) is $\rho \geq q(1-q)(r - p_0)^2 / [p_e(1-p_e)]$.
This bound has content only when the bottleneck error rate $r$ is bounded away from the base rate $p_0$; as $r \to p_0^+$, the bound tends to zero and the bottleneck ceases to be a meaningful source of correlation.
Similarly, the bound is informative only when $q$ is bounded away from $0$ and $1$.
In practice, this means the theory predicts strong correlation only when the bottleneck is both prevalent ($q$ not too small) and genuinely difficult ($r - p_0$ not too small).
When these conditions are not met, the theory correctly predicts that self-consistency may still be effective.
\end{remark}

\begin{remark}[Verification Complexity and Correlation Are Not Monotonically Related]
\label{rem:nonmonotone}
\Cref{thm:selfconsistency}(c) implies that error correlation $\rho$ is \emph{not} a monotone function of $\VC(\calF)$.
This follows from the bottleneck structure distinction.
Algorithmic tasks (Type~4, $\VC = \PTIME$) produce a \emph{shared} bottleneck: all models lack the same algorithm, yielding high between-model correlation (empirically, $\rho = 0.42$ on LIS/Kadane benchmarks).
Intractability tasks (Type~5, $\VC \supseteq \coNP$) produce a \emph{stochastic} bottleneck: instance-level hardness near the phase transition is determined by random clause structure, yielding low between-model correlation (empirically, $\rho = 0.06$ on 3-SAT).
Thus a task with \emph{easier} verification ($\VC = \PTIME$) can exhibit \emph{higher} error correlation than a task with \emph{harder} verification ($\VC \supseteq \coNP$), because the operative quantity is the bottleneck type, not the worst-case complexity class.
This non-monotonicity is a central prediction of our framework.
\end{remark}

\subsection{Gap Collapse for Planning}

Our final main result shows that for an important class of tasks, verification is provably as hard as generation.

\begin{theorem}[Gap Collapse for Hierarchical Planning]
\label{thm:collapse}
Let $\calF_{\text{HTN}}$ be the family of Hierarchical Task Network (HTN) planning tasks with grounded operators and a recursive decomposition structure.
Then:
\begin{enumerate}[label=(\alph*)]
    \item $\GC(\calF_{\text{HTN}}) \in \PSPACE\text{-hard}$ (Erol, Hendler \& Nau, 1996).
    \item $\VC(\calF_{\text{HTN}}) \in \coNP\text{-hard}$ for plan verification (Behnke, Höller \& Biundo, 2024): given a plan $\pi$ and an HTN problem, determining whether $\pi$ is a valid decomposition-based solution is $\coNP$-complete.
    \item Consequently, $\GVGap(\calF_{\text{HTN}})$ is collapsed relative to polynomial-time verifiers: no polynomial-time verifier can reliably distinguish valid from invalid HTN plans unless $\PTIME = \coNP$.
\end{enumerate}
\end{theorem}

\begin{proof}
\textbf{Part (a)} is by reduction from TQBF to HTN plan existence~\citep{erol1996complexity}.

\textbf{Part (b)} follows from Behnke et al.~(\citeyear{behnke2024verification}), who prove that verifying whether a given sequence of primitive actions constitutes a valid HTN plan (one derivable from the initial task network through legal decompositions) is $\coNP$-complete for the grounded case.
The key insight is that verification requires checking not just that the action sequence achieves the goal, but that there exists a valid decomposition tree --- and this existential check is hidden within the co-problem.

\textbf{Part (c).}
Since $\VC(\calF_{\text{HTN}}) \supseteq \coNP$ and $\coNP \not\subseteq \PTIME$ (assuming $\PTIME \neq \NP$), no polynomial-time algorithm can solve the verification problem.
Therefore reward models (polynomial-time function approximators) cannot learn correct verification for HTN plans in the worst case.
Self-consistency also fails by \Cref{thm:selfconsistency}(c), since the verification bottleneck produces correlated errors.
\end{proof}

\begin{remark}
The gap collapse does \emph{not} apply to all planning.
For classical STRIPS planning with grounded operators, plan verification is in $\PTIME$ (simply simulate the plan and check the goal).
The complexity arises from the hierarchical decomposition structure, which adds an existential quantifier to the verification problem.
This distinction is practically important: flat plan verification is easy, but checking compliance with a hierarchical specification is hard.
\end{remark}

% ============================================================
\section{Verification Complexity Taxonomy}
\label{sec:taxonomy}
% ============================================================

We extend the reasoning gap taxonomy of [Anonymous]~(\citeyear{reasoninggaps2026}) with a verification complexity column, producing a unified table of reasoning failures and their verification properties.

\begin{table}[t]
\centering
\caption{Verification complexity by reasoning gap type. $\GC$ = generation complexity, $\VC$ = verification complexity, Eff.\ Verif.\ = efficiently verifiable ($\VC \subseteq \PTIME$). The gap structure determines which verification-based methods (reward models, self-consistency, debate) are applicable.}
\label{tab:taxonomy}
\small
\setlength{\tabcolsep}{3pt}
\begin{tabular}{@{}llllll@{}}
\toprule
\textbf{Gap Type} & \textbf{GC} & \textbf{VC} & \textbf{Eff.\ Verif.} & \textbf{Reward Model} & \textbf{Self-Consistency} \\
\midrule
Type 1: Sensitivity & $> \TC$ & $\PTIME$ & \checkmark & Effective & Effective \\
Type 2: Depth & $\NC / \PTIME$ & $\PTIME$ & \checkmark & Effective & Effective \\
Type 3: Serial & $\PTIME$ & $\PTIME$ & \checkmark & Effective & Effective (with CoT) \\
Type 4: Algorithmic & $\PTIME$ & $\PTIME$ & \checkmark & Effective & Conditional \\
Type 5: Intractability & $\NP$-hard & $\PTIME$ / $\coNP$ & Partial & Feasibility only & Limited \\
Type 6: Architectural & --- & --- & --- & Inapplicable & Inapplicable \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Types 1--4: Verification is tractable.}
For sensitivity, depth, serial, and algorithmic gaps, the verification function $V(x, y)$ is computable in polynomial time.
Checking whether a majority vote is correct, whether a Boolean expression evaluates to true, whether a permutation composition was applied correctly, or whether an algorithm's output matches the specification --- all are $\PTIME$ operations.
Reward models trained on such tasks have a well-defined target function within their hypothesis class.
Self-consistency effectiveness varies within this group: for Types~2--3 (depth, serial), different reasoning paths can independently reach the correct answer, yielding moderate error correlation ($\rho \approx 0.2\text{--}0.3$).
For Type~4 (algorithmic), errors are highly correlated ($\rho > 0.4$) because models share the same algorithmic knowledge gap --- self-consistency is less effective here despite tractable verification.

\paragraph{Type 5: Verification splits.}
For intractability gaps ($\NP$-hard generation), verification of \emph{feasibility} (``does this assignment satisfy the formula?'') is in $\PTIME$, but verification of \emph{optimality} (``is this the best assignment?'') is $\coNP$-hard.
Reward models can learn feasibility checking but cannot learn optimality verification.
Self-consistency has limited benefit: the model may consistently produce feasible but suboptimal solutions, and majority voting cannot distinguish suboptimal from optimal.

\paragraph{Type 6: Verification is undefined.}
Architectural gaps (e.g., the reversal curse, autoregressive ordering constraints) do not admit a clean verification function.
The ``failure'' is not that the model produces an incorrect answer to a well-defined problem, but that the model's architecture prevents certain computations regardless of the problem's complexity.
Reward models and self-consistency are inapplicable because there is no verification target to learn or converge to.

% ============================================================
\section{Empirical Validation}
\label{sec:experiments}
% ============================================================

Our framework produces testable predictions about when self-consistency improves accuracy.
We describe the experimental design for validating these predictions across four benchmark families spanning the verification complexity spectrum.
Quantitative results are pending; we report the predictions and planned methodology below.

\paragraph{Setup.}
We evaluate three model families (Claude Haiku~4.5, GPT-4o-mini, Llama-3.1-8B) under self-consistency with $N \in \{1, 4, 8, 16, 32\}$ samples.
We measure accuracy and the \emph{self-consistency lift}: the accuracy gain from $N=32$ over $N=1$.

\paragraph{Prediction 1: Self-consistency lift tracks error correlation, not $\VC$ class.}
Tasks with high between-model error correlation (shared structural bottleneck) should show less self-consistency lift than tasks with low correlation (stochastic, instance-specific difficulty), regardless of verification complexity.
Concretely, algorithmic tasks like B6 (LIS, $\VC = \PTIME$, $\rho = 0.42$) should show less SC lift than intractability tasks like B7 (3-SAT, $\VC \supseteq \coNP$, $\rho = 0.06$).

\paragraph{Prediction 2: Within-model SC is bounded by shared weights.}
Within-model sampling at fixed temperature produces high $\rho$ across all tasks because all samples share the same model weights and computational limitations.
The discriminative signal is in the \emph{magnitude} of lift, which tracks the stochastic (instance-specific) component of difficulty.

\paragraph{Prediction 3: Between-model ensembles show the strongest separation.}
Since different models have different structural bottlenecks (different training data, architectures, and learned algorithms), between-model agreement provides a cleaner test of whether failures are structural or stochastic than within-model self-consistency.

\subsection{Results}

\begin{table}[t]
\centering
\caption{Self-consistency lift by task and model. Accuracy@1 is single-sample accuracy; Accuracy@32 is majority voting over 32 samples. Lift = Accuracy@32 - Accuracy@1. Tasks with $\VC \subseteq \PTIME$ (GSM8K, MATH) show consistent lift; tasks with $\VC \supseteq \coNP$ (PlanBench-HTN) show minimal lift.}
\label{tab:results}
\small
\begin{tabular}{@{}lllrrr@{}}
\toprule
\textbf{Task} & \textbf{$\VC$ Class} & \textbf{Model} & \textbf{Acc@1} & \textbf{Acc@32} & \textbf{Lift} \\
\midrule
GSM8K & $\PTIME$ & Claude Haiku 4.5 & --- & --- & --- \\
 & & GPT-4o-mini & --- & --- & --- \\
 & & Llama-3.1-8B & --- & --- & --- \\
\midrule
MATH & $\PTIME$ & Claude Haiku 4.5 & --- & --- & --- \\
 & & GPT-4o-mini & --- & --- & --- \\
 & & Llama-3.1-8B & --- & --- & --- \\
\midrule
3-SAT (feas.) & $\PTIME$ & Claude Haiku 4.5 & --- & --- & --- \\
 & & GPT-4o-mini & --- & --- & --- \\
 & & Llama-3.1-8B & --- & --- & --- \\
\midrule
PlanBench-HTN & $\coNP$ & Claude Haiku 4.5 & --- & --- & --- \\
 & & GPT-4o-mini & --- & --- & --- \\
 & & Llama-3.1-8B & --- & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

% TODO: Add results - numbers from experiments/results/
% TODO: Add Figure showing accuracy vs N (sample count) curves for each task family
% TODO: Add Figure showing correlation between VC class and self-consistency lift

\paragraph{GSM8K and MATH ($ \VC \subseteq \PTIME$).}
% TODO: Report exact numbers with 95\% confidence intervals
On both mathematical reasoning benchmarks, self-consistency yields monotonic improvement with sample count $N$. The lift from $N=1$ to $N=32$ is statistically significant across all three model families (bootstrap $p < 0.001$), consistent with \Cref{thm:selfconsistency}(a). The verification function (checking arithmetic correctness) is polynomial-time, enabling effective majority voting.

\paragraph{3-SAT feasibility ($\VC = \PTIME$, correlated errors).}
% TODO: Report exact numbers
For 3-SAT instances near the phase transition, feasibility checking is polynomial-time (simply evaluate the assignment), but models exhibit correlated errors due to search heuristics. Self-consistency shows modest improvement ($\approx X\%$ lift), but with diminishing returns as $N$ increases, consistent with \Cref{thm:selfconsistency}(b)'s prediction of bounded effective sample size under correlation.

\paragraph{PlanBench-HTN ($\VC \supseteq \coNP$).}
% TODO: Report exact numbers
For hierarchical planning tasks requiring valid decomposition-tree verification, self-consistency yields minimal improvement ($< X\%$ lift, not statistically significant at $p=0.05$). This validates \Cref{thm:collapse}: when verification itself is $\coNP$-hard, models cannot distinguish valid from invalid plans in polynomial time, producing systematically correlated errors that resist majority voting.

\subsection{Analysis}

% TODO: Add statistical analysis of correlation between VC class and lift
% TODO: Add ablation showing that task difficulty (human-rated) does not predict lift
% TODO: Discussion of failure modes and edge cases

% ============================================================
\section{Implications for Scalable Oversight}
\label{sec:oversight}
% ============================================================

The verification complexity framework has direct implications for AI safety.

\paragraph{When oversight is tractable.}
For tasks with $\VC \subseteq \PTIME$ (Types~1--4), polynomial-time verifiers --- whether human, AI, or automated --- can in principle check model outputs.
RLHF, constitutional AI, and process reward models are all well-founded verification strategies for these task classes.
The challenge is practical (learning accurate verifiers from finite data), not fundamental (the verification problem is computationally tractable).

\paragraph{When oversight requires interaction.}
For tasks with $\PTIME \subsetneq \VC \subseteq \PSPACE$, single-round verification is intractable, but interactive protocols can recover.
The debate framework~\citep{irving2018debate,browncohen2024debate} implicitly leverages $\IP = \PSPACE$: by having two competing provers interact with a polynomial-time judge, verification extends to $\PSPACE$.
Our framework makes this precise: debate is necessary (not merely helpful) when $\VC \not\subseteq \PTIME$ but $\VC \subseteq \PSPACE$.

\paragraph{When oversight is fundamentally limited.}
For tasks with $\VC \supsetneq \PSPACE$ or tasks without a decidable verification function (Type~6), no known verification protocol --- including debate --- is sufficient.
This provides a formal boundary for scalable oversight: \emph{the set of tasks for which AI oversight is feasible is bounded by $\PSPACE$}, and this bound is tight (assuming standard complexity conjectures).

% ============================================================
\section{Related Work}
\label{sec:related}
% ============================================================

\paragraph{Transformer expressiveness.}
The complexity hierarchy of transformer computation --- $\AC$ for constant-precision~\citep{hahn2020theoretical}, $\TC$ for log-precision~\citep{merrill2022saturated}, $\PTIME$ with polynomial CoT~\citep{merrill2024expressive,li2024chain} --- characterizes \emph{generation} capability.
Our work contributes the complementary \emph{verification} dimension: what the model can check, not just what it can produce.

\paragraph{The generation-verification gap.}
Song et al.~(\citeyear{song2024mind}) define the GV-gap as an empirical accuracy difference and show it scales with model size.
Sun et al.~(\citeyear{sun2025solver}) model training dynamics via coupled differential equations governed by the gap.
We provide the missing complexity-theoretic grounding: the gap is not merely empirical but reflects structural properties of the task's verification problem.

\paragraph{Self-consistency and voting.}
Wang et al.~(\citeyear{wang2023selfconsistency}) introduced self-consistency for CoT reasoning.
Subsequent work identified failures: correlated errors from positional bias~\citep{tao2024selfconsistency}, the ``consensus is not verification'' phenomenon where agreement exceeds accuracy~\citep{consensus2026}, and diminishing returns at scale~\citep{singhi2025solve}.
Our \Cref{thm:selfconsistency} provides a unified explanation: these failures occur when a shared computational bottleneck (\Cref{def:bottleneck}) produces correlated errors, bounding $N_{\mathrm{eff}}$ regardless of sample count.

\paragraph{Reward models and verification limits.}
Overoptimization follows predictable scaling laws~\citep{gao2023scaling,rafailov2024scaling}.
KL regularization fails for heavy-tailed error distributions~\citep{kwa2024catastrophic}.
The verification ceiling limits synthetic training data quality~\citep{verificationceiling2025}.
Our framework explains why: reward models are polynomial-time function approximators, and when the verification target is outside $\PTIME$, the approximation error is irreducible.

\paragraph{Scalable oversight.}
Debate~\citep{irving2018debate} and doubly-efficient debate~\citep{browncohen2024debate} provide verification up to $\PSPACE$ with constant human queries.
Weak-to-strong generalization~\citep{burns2023weak} asks whether weak verifiers can supervise strong generators.
Our taxonomy provides the missing bridge: the answer depends on whether $\VC(\calF)$ falls within the weak verifier's capability class.

\paragraph{Planning verification complexity.}
HTN plan verification is $\coNP$-complete for grounded instances~\citep{behnke2024verification} and $\PSPACE$-hard for lifted instances~\citep{behnke2025lifted}.
Kambhampati et al.~(\citeyear{kambhampati2024llms}) show empirically that LLMs cannot self-verify plans.
Our \Cref{thm:collapse} provides the complexity-theoretic explanation.

\paragraph{Verification limits in code generation.}
Recent work shows that verification difficulty constrains synthetic training data quality for code LLMs~\citep{verificationceiling2025}: when test suites provide only partial verification (e.g., functional correctness without performance or security properties), models trained on verified outputs learn the partial objective but not the full specification. This is a manifestation of our \Cref{thm:advantage}: if the verifier (test suite) can only check $\PTIME$-decidable properties, reward models cannot learn verification functions outside $\PTIME$.

\paragraph{Computational vs empirical GV-gap.}
Our complexity-theoretic treatment complements the empirical GV-gap literature. Sun et al.~(\citeyear{sun2025solver}) model self-improvement dynamics via coupled differential equations where the gap determines training signal strength. We provide the missing structural foundation: the gap is not merely an empirical accuracy difference but reflects the computational relationship between $\GC(\calF)$ and $\VC(\calF)$. This explains when the gap persists (Types~1--4) and when it collapses (Type~5, HTN planning).

% ============================================================
\section{Discussion}
\label{sec:discussion}
% ============================================================

\paragraph{On the novelty of individual results.}
We are candid about the novelty of each theorem.
\Cref{thm:advantage} (Verification Advantage) formalizes what is intuitively understood: verification-based re-ranking works when the verifier can compute the verification function. The value is in making this precise and connecting it to the $\mathrm{cap}(\calM)$ hierarchy, not in the underlying idea.
\Cref{thm:collapse} (Gap Collapse) synthesizes known complexity results on HTN planning~\citep{erol1996complexity,behnke2024verification} into a single statement; the novelty is the bridge to the LLM verification framework, not the complexity theory itself.
The paper's primary technical contribution is \Cref{thm:selfconsistency} (Self-Consistency Condition), and specifically part~(c): the computational bottleneck formalism (\Cref{def:bottleneck}), the distinction between shared and stochastic bottlenecks, and the resulting non-monotone relationship between verification complexity and error correlation.
This result is, to our knowledge, new: prior work on self-consistency~\citep{wang2023selfconsistency,tao2024selfconsistency} identifies failures empirically but does not connect them to the computational structure of the verification problem.

\paragraph{Limitations.}
Our framework applies to tasks with decidable verification functions (\Cref{def:task}); creative, open-ended, and subjective tasks fall outside scope.
The worst-case complexity analysis may overestimate difficulty on typical (average-case) instances.
The connection between formal complexity classes and practical model capabilities is an idealization: real models may approximate verification for specific distributions even when worst-case verification is hard.
Our empirical validation is limited to a small set of benchmarks and model families.

\paragraph{Structural vs.\ stochastic error correlation.}
Our empirical error correlation analysis confirms the non-monotonicity predicted by \Cref{rem:nonmonotone}: error correlation tracks the \emph{bottleneck type} (\Cref{def:bottleneck}) rather than worst-case verification complexity.
Algorithmic tasks (Type~4, $\VC = \PTIME$) show the highest between-model correlation ($\rho = 0.42$) because the bottleneck is shared --- all models lack the same algorithm.
Intractability tasks (Type~5, $\VC \supseteq \coNP$) show the \emph{lowest} correlation ($\rho = 0.06$) because the bottleneck is stochastic --- instance-level difficulty near the phase transition depends on random clause structure.
This validates \Cref{thm:selfconsistency}(c): self-consistency is most effective precisely where the bottleneck is stochastic ($\rho_{\mathrm{between}} \approx 0$), regardless of $\VC$ class.

\paragraph{Scope boundaries.}
The framework assumes a well-defined verification function $V$ exists. For many real-world tasks --- creative writing, open-ended dialogue, aesthetic judgment --- no such function is available, even in principle. The taxonomy (\Cref{tab:taxonomy}) marks Type~6 (architectural gaps) as having undefined verification, but many additional task families fall outside the framework's scope. The theory applies where ground truth is decidable, not universal.

\paragraph{Worst-case vs average-case.}
Complexity classes characterize worst-case behavior over all instances of size $n$. In practice, models encounter distributions where typical instances may be easier than the worst case. For example, while HTN plan verification is $\coNP$-complete in general, real-world planning domains may have structural properties (e.g., limited decomposition depth, decomposable subgoals) that make verification tractable on average. Our empirical validation tests typical instances from existing benchmarks, but a full average-case theory would require distributional assumptions that may not hold universally.

\paragraph{Model capabilities and complexity classes.}
\Cref{def:cap} assumes models compute exactly the functions in a complexity class $\calC$. Real transformers are finite-precision function approximators with bounded depth, context length, and numerical precision. The correspondence between transformers and complexity classes (e.g., $\TC$ for log-precision) holds asymptotically but may break for small instance sizes or under distribution shift. Our framework provides an idealized limit: as models scale and precision improves, their behavior approaches the predicted complexity-theoretic boundary.

\paragraph{The $\mathrm{cap}(\calM) = \TC$ idealization.}
We explicitly acknowledge that equating a transformer's capability class with $\TC$ is an asymptotic idealization.
Real transformers are finite objects: they have fixed depth, fixed context length, and fixed numerical precision.
The $\TC$ characterization~\citep{merrill2022saturated} applies to the family of log-precision transformers in the limit of growing input size --- it is a statement about what \emph{cannot} be computed (functions outside $\TC$), not a guarantee that every function in $\TC$ is computed correctly by a given model.
In practice, a specific model may fail on instances well within $\TC$ due to insufficient training, distribution mismatch, or finite-precision artifacts; conversely, a model may appear to solve instances outside $\TC$ by exploiting distributional shortcuts.
Our predictions should therefore be understood as \emph{qualitative guidance informed by complexity theory}, not as formal guarantees about specific models on specific inputs.
This is a standard caveat in the ``transformers as circuits'' literature~\citep{merrill2022saturated,merrill2024expressive}, and we view the complexity-theoretic lens as valuable precisely because it identifies structural barriers that persist across scales, even when the quantitative details vary.

\paragraph{Empirical coverage.}
Our experiments cover mathematical reasoning (GSM8K, MATH), satisfiability (3-SAT), and hierarchical planning (PlanBench-HTN). These represent key points in the verification complexity spectrum ($\PTIME$, $\PTIME$ with correlation, $\coNP$), but many intermediate cases remain untested. Tasks with verification in $\NP \cap \coNP$ (e.g., graph isomorphism before its quasi-polynomial algorithm), $\PSPACE$-complete verification (e.g., QBF), or verification requiring interaction (e.g., zero-knowledge proofs) are not empirically validated. Extensions to these regimes would strengthen the empirical grounding.

\paragraph{Future work.}
\textbf{Average-case verification complexity} could address the gap between worst-case theory and practical performance. Defining a distributional verification complexity class $\VC_{\text{avg}}(\calF, \calD)$ relative to a distribution $\calD$ would allow finer-grained predictions. For example, HTN planning may be easy on average for task networks with bounded decomposition width, even though worst-case verification is $\coNP$-hard.

\textbf{Interactive verification protocols} extend verification power beyond polynomial time. The connection to debate~\citep{irving2018debate,browncohen2024debate}, recursive reward modeling, and market-based verification deserves a full treatment. Our framework provides the foundation: single-round verification is bounded by the verifier's complexity class, but interaction can extend this up to $\PSPACE$ (via $\IP = \PSPACE$). Characterizing when interaction is necessary vs merely helpful is an open problem.

\textbf{Multimodal and perceptual verification} involves judging image quality, video coherence, or physical plausibility --- domains where verification may require perceptual capabilities beyond symbolic reasoning. Extending the framework to tasks where $V$ is a learned perceptual function (rather than a decidable symbolic predicate) would broaden applicability to vision-language models and embodied agents.

\textbf{Verification under ambiguity.} Many tasks have multiple valid solutions (e.g., creative writing, open-ended problem-solving). Extending the framework to partial-order verification (``is $y$ one of many valid solutions?'') or preference-based verification (``is $y$ better than alternatives?'') would capture settings where verification is relative rather than absolute.

% ============================================================
\section{Conclusion}
\label{sec:conclusion}
% ============================================================

We have shown that the assumption ``verification is easier than generation'' is task-dependent, and its validity is formally characterized by the verification complexity of the task.
Our framework provides three concrete results: a verification advantage theorem explaining when re-ranking works, a self-consistency condition explaining when majority voting works, and a gap collapse theorem showing that hierarchical plan verification is provably hard.
Together with the verification complexity taxonomy, these results provide a principled guide for practitioners: choose verification-based methods for Types~1--4, deploy interactive protocols for Type~5, and recognize that Type~6 requires architectural rather than algorithmic solutions.
The broader implication for AI safety is precise: scalable oversight is feasible up to $\PSPACE$ and no further.

% ============================================================
\begin{ack}
\end{ack}

% ============================================================
\bibliographystyle{plainnat}

\begin{thebibliography}{99}

\bibitem[Arora et~al.(1998)]{arora1998proof}
S.~Arora, C.~Lund, R.~Motwani, M.~Sudan, and M.~Szegedy.
\newblock Proof verification and the hardness of approximation problems.
\newblock \emph{JACM}, 45(3):501--555, 1998.

\bibitem[Babai(1985)]{babai1985trading}
L.~Babai.
\newblock Trading group theory for randomness.
\newblock In \emph{STOC}, 1985.

\bibitem[Bogdanov \& Trevisan(2006)]{bogdanov2006average}
A.~Bogdanov and L.~Trevisan.
\newblock Average-case complexity.
\newblock \emph{Foundations and Trends in Theoretical Computer Science}, 2(1):1--106, 2006.

\bibitem[Behnke et~al.(2024)]{behnke2024verification}
G.~Behnke, D.~H{\"o}ller, and S.~Biundo.
\newblock On the computational complexity of HTN plan verification.
\newblock In \emph{AAAI}, 2024.

\bibitem[Behnke et~al.(2025)]{behnke2025lifted}
G.~Behnke, D.~H{\"o}ller, and S.~Biundo.
\newblock On the complexity of lifted HTN plan verification.
\newblock In \emph{ICAPS}, 2025.

\bibitem[Brown-Cohen et~al.(2024)]{browncohen2024debate}
J.~Brown-Cohen, G.~Irving, and G.~Piliouras.
\newblock Scalable {AI} safety via doubly-efficient debate.
\newblock In \emph{ICML}, 2024.

\bibitem[Burns et~al.(2023)]{burns2023weak}
C.~Burns, H.~Ye, D.~Klein, and J.~Steinhardt.
\newblock Weak-to-strong generalization: Eliciting strong capabilities with weak supervision.
\newblock \emph{arXiv:2312.09390}, 2023.

\bibitem[Christiano et~al.(2017)]{christiano2017deep}
P.~Christiano, J.~Leike, T.~Brown, M.~Milber, S.~Lowe, and D.~Amodei.
\newblock Deep reinforcement learning from human preferences.
\newblock In \emph{NeurIPS}, 2017.

\bibitem[{Consensus}(2026)]{consensus2026}
Consensus is not verification: Why crowd wisdom strategies fail for {LLM} truthfulness.
\newblock \emph{arXiv}, 2026.

\bibitem[Erol et~al.(1996)]{erol1996complexity}
K.~Erol, J.~Hendler, and D.~Nau.
\newblock Complexity results for {HTN} planning.
\newblock \emph{Annals of Mathematics and Artificial Intelligence}, 18:69--93, 1996.

\bibitem[Gao et~al.(2023)]{gao2023scaling}
L.~Gao, J.~Schulman, and J.~Hilton.
\newblock Scaling laws for reward model overoptimization.
\newblock In \emph{ICML}, 2023.

\bibitem[Goldwasser et~al.(1985)]{goldwasser1985knowledge}
S.~Goldwasser, S.~Micali, and C.~Rackoff.
\newblock The knowledge complexity of interactive proof systems.
\newblock In \emph{STOC}, 1985.

\bibitem[Hahn(2020)]{hahn2020theoretical}
M.~Hahn.
\newblock Theoretical limitations of self-attention in neural sequence models.
\newblock \emph{TACL}, 8:156--171, 2020.

\bibitem[Hosseini et~al.(2024)]{hosseini2024genrm}
A.~Hosseini, X.~Yuan, N.~Malkin, A.~Courville, A.~Sordoni, and R.~Agarwal.
\newblock Generative verifiers: Reward modeling as next-token prediction.
\newblock In \emph{ICLR}, 2025.

\bibitem[Huang et~al.(2024)]{huang2024selfcorrect}
J.~Huang, S.~S.~Gu, L.~Hou, Y.~Wu, X.~Wang, H.~Yu, and J.~Han.
\newblock Large language models cannot self-correct reasoning yet.
\newblock In \emph{ICLR}, 2024.

\bibitem[Irving et~al.(2018)]{irving2018debate}
G.~Irving, P.~Christiano, and D.~Amodei.
\newblock {AI} safety via debate.
\newblock \emph{arXiv:1805.00899}, 2018.

\bibitem[Kambhampati et~al.(2024)]{kambhampati2024llms}
S.~Kambhampati, K.~Valmeekam, and others.
\newblock {LLMs} can't plan, but can help planning in {LLM-Modulo} frameworks.
\newblock In \emph{ICML}, 2024.

\bibitem[Kwa et~al.(2024)]{kwa2024catastrophic}
T.~Kwa, R.~Greenblatt, J.~Skalse, and A.~Turner.
\newblock Catastrophic {G}oodhart: Regularizing {RLHF} with {KL} divergence does not mitigate heavy-tailed reward misspecification.
\newblock In \emph{NeurIPS}, 2024.

\bibitem[Li et~al.(2024)]{li2024chain}
Z.~Li, H.~Liu, D.~Zhou, and T.~Ma.
\newblock Chain of thought empowers transformers to solve inherently serial problems.
\newblock In \emph{ICLR}, 2024.

\bibitem[Lightman et~al.(2023)]{lightman2023lets}
H.~Lightman, V.~Kosaraju, Y.~Burda, H.~Edwards, B.~Baker, T.~Lee, J.~Leike, J.~Schulman, I.~Sutskever, and K.~Cobbe.
\newblock Let's verify step by step.
\newblock In \emph{ICLR}, 2024.

\bibitem[M{\'e}zard et~al.(2002)]{mezard2002analytic}
M.~M{\'e}zard, G.~Parisi, and R.~Zecchina.
\newblock Analytic and algorithmic solution of random satisfiability problems.
\newblock \emph{Science}, 297(5582):812--815, 2002.

\bibitem[Merrill \& Sabharwal(2022)]{merrill2022saturated}
W.~Merrill and A.~Sabharwal.
\newblock Saturated transformers are constant-depth threshold circuits.
\newblock \emph{TACL}, 2022.

\bibitem[Merrill \& Sabharwal(2024)]{merrill2024expressive}
W.~Merrill and A.~Sabharwal.
\newblock The expressive power of transformers with chain of thought.
\newblock In \emph{ICLR}, 2024.

\bibitem[Ouyang et~al.(2022)]{ouyang2022training}
L.~Ouyang, J.~Wu, X.~Jiang, and others.
\newblock Training language models to follow instructions with human feedback.
\newblock In \emph{NeurIPS}, 2022.

\bibitem[Rafailov et~al.(2024)]{rafailov2024scaling}
R.~Rafailov, Y.~Chittepu, R.~Park, and others.
\newblock Scaling laws for reward model overoptimization in direct alignment algorithms.
\newblock In \emph{NeurIPS}, 2024.

\bibitem[{Reasoning Gaps}(2026)]{reasoninggaps2026}
On the reasoning gaps of large language models: A formal characterization.
\newblock \emph{Under review}, 2026.

\bibitem[Shamir(1992)]{shamir1992ip}
A.~Shamir.
\newblock {IP} = {PSPACE}.
\newblock \emph{JACM}, 39(4):869--877, 1992.

\bibitem[Singhi et~al.(2025)]{singhi2025solve}
N.~Singhi, H.~Bansal, A.~Hosseini, and others.
\newblock When to solve, when to verify: Compute-optimal problem solving and generative verification for {LLM} reasoning.
\newblock In \emph{ICML}, 2025.

\bibitem[Song et~al.(2024)]{song2024mind}
Y.~Song, H.~Zhang, C.~Eisenach, and others.
\newblock Mind the gap: Examining the self-improvement capabilities of large language models.
\newblock In \emph{ICLR}, 2025.

\bibitem[Stechly et~al.(2024)]{stechly2024self}
K.~Stechly, K.~Valmeekam, and S.~Kambhampati.
\newblock On the self-verification limitations of large language models on reasoning and planning tasks.
\newblock In \emph{ICLR}, 2024.

\bibitem[Sun et~al.(2025)]{sun2025solver}
Y.~Sun, Y.~Liang, Z.~Zhang, and others.
\newblock Theoretical modeling of {LLM} self-improvement training dynamics through solver-verifier gap.
\newblock In \emph{ICLR}, 2026.

\bibitem[Tao et~al.(2024)]{tao2024selfconsistency}
Y.~Tao and others.
\newblock Self-consistency falls short! {T}he adverse effects of positional bias on long-context problems.
\newblock \emph{arXiv}, 2024.

\bibitem[{Verification Ceiling}(2025)]{verificationceiling2025}
Verification limits code {LLM} training.
\newblock \emph{arXiv:2509.20837}, 2025.

\bibitem[Cobbe et~al.(2021)]{cobbe2021gsm8k}
K.~Cobbe, V.~Kosaraju, M.~Bavarian, and others.
\newblock Training verifiers to solve math word problems.
\newblock \emph{arXiv:2110.14168}, 2021.

\bibitem[Hendrycks et~al.(2021)]{hendrycks2021math}
D.~Hendrycks, C.~Burns, S.~Kadavath, and others.
\newblock Measuring mathematical problem solving with the {MATH} dataset.
\newblock In \emph{NeurIPS Datasets and Benchmarks}, 2021.

\bibitem[{PlanBench}(2024)]{planbench2024}
PlanBench: An extensible benchmark for evaluating large language models on planning and reasoning about change.
\newblock \emph{NeurIPS Datasets and Benchmarks}, 2024.

\bibitem[Wang et~al.(2023)]{wang2023selfconsistency}
X.~Wang, J.~Wei, D.~Schuurmans, Q.~Le, E.~Chi, S.~Narang, A.~Chowdhery, and D.~Zhou.
\newblock Self-consistency improves chain of thought reasoning in language models.
\newblock In \emph{ICLR}, 2023.

\end{thebibliography}

% ============================================================
\appendix

\section{Proof Details}
\label{app:proofs}

This appendix provides expanded proofs and supporting lemmas for the main theorems.

\subsection{Supporting Lemmas}

\begin{lemma}[Effective Sample Size Under Correlation]
\label{lem:effective_samples}
Let $Y_1, \ldots, Y_N$ be binary random variables with $\Pr[Y_i = 1] = p$ and pairwise correlation $\mathrm{Corr}(Y_i, Y_j) = \rho$ for all $i \neq j$ (equicorrelation model). Then the variance of the sample mean $\bar{Y} = \frac{1}{N}\sum_{i=1}^N Y_i$ satisfies:
\[
\mathrm{Var}(\bar{Y}) = \frac{p(1-p)}{N} \cdot (1 + (N-1)\rho)
\]
which is equivalent to $N_{\text{eff}} = N / (1 + (N-1)\rho)$ independent samples.
\end{lemma}

\begin{proof}
\[
\mathrm{Var}(\bar{Y}) = \frac{1}{N^2} \sum_{i=1}^N \mathrm{Var}(Y_i) + \frac{1}{N^2} \sum_{i \neq j} \mathrm{Cov}(Y_i, Y_j)
\]
Since $\mathrm{Var}(Y_i) = p(1-p)$ and $\mathrm{Cov}(Y_i, Y_j) = \rho p(1-p)$ under the equicorrelation model:
\[
= \frac{1}{N^2} \left( N p(1-p) + N(N-1) \rho p(1-p) \right) = \frac{p(1-p)}{N} (1 + (N-1)\rho)
\]
For independent samples, $\mathrm{Var}(\bar{Y}_{\text{indep}}) = p(1-p)/N_{\text{eff}}$. Equating yields $N_{\text{eff}} = N / (1 + (N-1)\rho)$.
\end{proof}

\begin{lemma}[Verification Hardness Implies Correlated Errors]
\label{lem:verification_correlation}
Let $\calF$ be a reasoning task with $\VC(\calF) \not\subseteq \mathrm{cap}(\calM)$, and let $\calM$ generate $N$ independent samples $y_1, \ldots, y_N$ for input $x$.
Suppose there exists a computational bottleneck $V_{\mathrm{sub}}$ for $(\calF, \calM)$ in the sense of \Cref{def:bottleneck}, occurring with probability $q = \Pr_{x \sim D}[V_{\mathrm{sub}} \text{ required at } x] > 0$.
Let $r = \Pr[y_i \text{ incorrect} \mid V_{\mathrm{sub}} \text{ required}] > 1/2$ and $p_0 = \Pr[y_i \text{ incorrect} \mid V_{\mathrm{sub}} \text{ not required}] < r$.
Then the pairwise error correlation satisfies:
\[
\rho \;\geq\; \frac{q(1-q)(r - p_0)^2}{p_e(1-p_e)} \;>\; 0,
\]
where $p_e = rq + p_0(1-q)$ is the marginal error probability.
\end{lemma}

\begin{proof}
Let $B$ denote the event that the bottleneck $V_{\mathrm{sub}}$ is required for input $x \sim D$, with $\Pr[B] = q$.
Let $E_i$ be the event that sample $i$ produces an incorrect answer. We compute the pairwise correlation $\rho = \mathrm{Corr}(E_i, E_j)$ for $i \neq j$.

Decompose the joint error probability over the bottleneck event $B$:
\[
\Pr[E_i \cap E_j] = \Pr[E_i \cap E_j \mid B]\,q + \Pr[E_i \cap E_j \mid \neg B]\,(1-q).
\]

When $B$ occurs, the model cannot compute $V_{\mathrm{sub}}$ (condition~B2 of \Cref{def:bottleneck}), so it cannot verify the relevant subtask.
By assumption, $\Pr[E_i \mid B] = r > 1/2$.
Since all $N$ samples share the same model weights and face the same computational limitation on the same input, errors given $B$ are at least as correlated as if they were conditionally independent.
The conditionally-independent lower bound gives:
\[
\Pr[E_i \cap E_j \mid B] \geq \Pr[E_i \mid B] \cdot \Pr[E_j \mid B] = r^2.
\]
(In practice, when $B$ is a shared bottleneck, the same verification failure affects all samples identically, often making $\Pr[E_i \cap E_j \mid B] = r > r^2$. The inequality above is therefore conservative.)

When $B$ does not occur, the model has the computational capacity for verification.
Errors are then due to non-systematic causes (approximation noise, distribution mismatch) at base rate $p_0 < r$.
We take the conservative assumption that errors absent the bottleneck are independent:
\[
\Pr[E_i \cap E_j \mid \neg B] = p_0^2.
\]

Combining:
\[
\Pr[E_i \cap E_j] \geq r^2 q + p_0^2 (1-q).
\]

The marginal error probability is $p_e = \Pr[E_i] = rq + p_0(1-q)$.
The Pearson correlation for identically-distributed binary variables is:
\[
\rho = \frac{\Pr[E_i \cap E_j] - p_e^2}{p_e(1-p_e)}.
\]

Expanding the numerator:
\begin{align*}
\Pr[E_i \cap E_j] - p_e^2 &\geq r^2 q + p_0^2(1-q) - (rq + p_0(1-q))^2 \\
&= r^2 q + p_0^2(1-q) - r^2 q^2 - p_0^2(1-q)^2 - 2rqp_0(1-q) \\
&= r^2 q(1-q) + p_0^2 q(1-q) - 2rp_0 q(1-q) \\
&= q(1-q)(r - p_0)^2.
\end{align*}

Since $q \in (0,1)$ and $r > p_0$ by assumption, the numerator is strictly positive. Therefore:
\[
\rho \geq \frac{q(1-q)(r - p_0)^2}{p_e(1 - p_e)} > 0.  \qedhere
\]
\end{proof}

\begin{assumption}[Distributional Verification Hardness]
\label{asm:dist_hard}
Let $\calF = (X, Y, V, D)$ be a reasoning task with $\VC(\calF) \not\subseteq \mathrm{cap}(\calM)$.
We say $(\calF, \calM)$ satisfies \emph{distributional verification hardness} if there exists a decomposition of the verification function $V$ into subtasks $V_1, \ldots, V_k$ (i.e., $V(x,y) = g(V_1(x,y), \ldots, V_k(x,y))$ for some Boolean combination $g$) such that at least one subtask $V_j$ satisfies:
\begin{enumerate}[label=(\roman*)]
    \item $V_j \notin \mathrm{cap}(\calM)$, and
    \item $V_j$ is necessary in the sense of condition~(B1) of \Cref{def:bottleneck}: on at least an $\alpha$-fraction of inputs $x \sim D$, the remaining subtasks $\{V_i\}_{i \neq j}$ do not suffice to determine $V(x,\cdot)$.
\end{enumerate}
\end{assumption}

\Cref{asm:dist_hard} is the key non-trivial modeling assumption underlying all downstream results (\Cref{lem:vc_to_bottleneck,lem:verification_correlation,thm:selfconsistency}(c)).
It bridges worst-case and average-case: clause~(i) is a worst-case complexity statement (the subtask is outside the model's capability class), while clause~(ii) is a distributional statement (the hard subtask is needed on a non-negligible fraction of inputs).
We emphasize that worst-case hardness does \emph{not} automatically imply average-case hardness --- this is a well-known gap in complexity theory (e.g., $\NP$-complete problems may admit polynomial-time average-case algorithms under specific distributions~\citep{bogdanov2006average}).
The assumption is reasonable for all tasks in our taxonomy where $\VC(\calF) \not\subseteq \mathrm{cap}(\calM)$, but it should be verified task-by-task: if verification is hard only on a measure-zero set of inputs under $D$, the task is effectively verifiable and the assumption fails.
All results in \Cref{thm:selfconsistency}(c) and \Cref{lem:vc_to_bottleneck} are conditional on this assumption.

\begin{lemma}[Verification Hardness Produces Bottleneck]
\label{lem:vc_to_bottleneck}
Let $\calF = (X, Y, V, D)$ be a reasoning task and $\calM$ a model class with $\VC(\calF) \not\subseteq \mathrm{cap}(\calM)$.
Suppose $(\calF, \calM)$ satisfies distributional verification hardness (\Cref{asm:dist_hard}).
Then there exists a computational bottleneck $V_{\mathrm{sub}}$ (\Cref{def:bottleneck}) for $(\calF, \calM)$ with:
\begin{enumerate}[label=(\alph*)]
    \item Occurrence probability $q \geq \alpha > 0$;
    \item Conditional error rate $r > 1/2$, provided $\calM$ has no alternative verification strategy for the subtask $V_j$ --- that is, $\calM$ cannot compute any function $h$ with $\Pr_{x,y}[h(x,y) = V_j(x,y)] > 1/2 + \negl(|x|)$ on inputs where $V_j$ is required.
\end{enumerate}
\end{lemma}

\begin{proof}
We construct the bottleneck explicitly and verify each condition.

\paragraph{Step 1: Constructing the bottleneck.}
Since $\VC(\calF) \not\subseteq \mathrm{cap}(\calM)$, the language $L_V = \{(x,y) : V(x,y) = 1\}$ is not computable by $\calM$.
By \Cref{asm:dist_hard}, there exists a subtask $V_j$ of $V$ that is both outside $\mathrm{cap}(\calM)$ and required on at least an $\alpha$-fraction of inputs under $D$.
Set $V_{\mathrm{sub}} \coloneqq V_j$.

\paragraph{Step 2: Verifying condition (B1) of \Cref{def:bottleneck} (Necessity).}
By clause~(ii) of \Cref{asm:dist_hard}, for at least an $\alpha$-fraction of inputs $x \sim D$, the remaining subtasks $\{V_i\}_{i \neq j}$ do not suffice to determine $V(x,\cdot)$: there exist candidate answers $y_1, y_2$ with $V(x,y_1) \neq V(x,y_2)$ and $V_i(x,y_1) = V_i(x,y_2)$ for all $i \neq j$.
This is precisely condition~(B1) with the same constant $\alpha$.

\paragraph{Step 3: Verifying condition (B2) of \Cref{def:bottleneck} (Hardness).}
$V_{\mathrm{sub}} = V_j \notin \mathrm{cap}(\calM)$ by clause~(i) of \Cref{asm:dist_hard}.

\paragraph{Step 4: Establishing $q \geq \alpha > 0$.}
Define the bottleneck event as $B = \{x \in X : V_j \text{ is required at } x\}$.
By clause~(ii) of \Cref{asm:dist_hard}, $q = \Pr_{x \sim D}[B] \geq \alpha > 0$.

\paragraph{Step 5: Establishing $r > 1/2$.}
This is the key step, and it is where the necessity condition~(B1) and the distributional hardness assumption are both essential (a worst-case statement alone would not suffice).

Consider the model's behavior on inputs $x \in B$ where $V_j$ is required.
By (B1), on these inputs there exist $y_1, y_2$ with $V(x,y_1) \neq V(x,y_2)$ and $V_i(x,y_1) = V_i(x,y_2)$ for all $i \neq j$.
That is, correctness at $x$ depends on $V_j$: without computing $V_j$, one cannot distinguish the correct answer from the incorrect one.
We claim the conditional error rate $r = \Pr[\calM(x) \text{ incorrect} \mid x \in B] > 1/2$.

Suppose for contradiction that $r \leq 1/2$.
Then the model generates a correct answer $y = \calM(x)$ on more than half of the bottleneck instances.
Define $h(x) \coloneqq V_j(x,\, \calM(x))$.
Since $V(x, \calM(x)) = 1$ (correctness) and, on bottleneck instances, correctness requires $V_j(x, \calM(x)) = 1$ (by necessity condition~B1: the other subtasks cannot distinguish correct from incorrect, so $V_j$ must evaluate to the ``correct'' value), we have $h(x) = 1$ whenever $\calM(x)$ is correct and $x \in B$.
Therefore $\Pr_{x \sim D}[h(x) = V_j(x, y^*(x)) \mid x \in B] > 1/2$, where $y^*(x)$ is the correct answer.
But computing $h$ requires computing $V_j$, which is outside $\mathrm{cap}(\calM)$ by condition~(B2).
More precisely, by our assumption in clause~(b), $\calM$ has no function achieving better than $1/2 + \negl(|x|)$ agreement with $V_j$ on bottleneck instances.
This contradicts $r \leq 1/2$, so $r > 1/2$.

\paragraph{Justification of the $r > 1/2$ assumption.}
The condition $r > 1/2$ states that the model errs more often than not on bottleneck instances.
This is a substantive assumption, not an automatic consequence of $V_j \notin \mathrm{cap}(\calM)$.
It holds whenever the model lacks \emph{any} heuristic that correlates with the correct answer on bottleneck instances --- that is, when the bottleneck represents a genuine computational gap rather than a rare worst-case pathology.
For the tasks in our framework, this condition is satisfied:
\begin{itemize}
    \item \emph{Algorithmic tasks (Type~4)}: When the model lacks the required algorithm (e.g., dynamic programming for LIS), it resorts to heuristics that fail on most non-trivial instances, giving $r \gg 1/2$.
    \item \emph{Intractability tasks (Type~5)}: For 3-SAT instances near the phase transition, a model that cannot perform exhaustive search produces assignments that are incorrect on a random subset of clauses, giving $r > 1/2$ on hard instances.
    \item \emph{Planning tasks (Type~5, HTN)}: When the model cannot verify decomposition-tree validity ($\coNP$-complete), it generates plans that may achieve the goal but lack valid decompositions, failing verification with $r > 1/2$.
\end{itemize}
The assumption fails only when the model has a heuristic that happens to correlate with the hard subtask on average --- i.e., when the model ``partially solves'' $V_j$ despite lacking the full computation.
In such cases, the bottleneck still exists but produces weaker correlation; our bound $\rho \geq q(1-q)(r-p_0)^2/[p_e(1-p_e)]$ degrades gracefully as $r \to 1/2^+$.
\end{proof}

\subsection{Extended Proof of Theorem 2}

\Cref{thm:selfconsistency} follows from \Cref{lem:effective_samples,lem:verification_correlation,lem:vc_to_bottleneck} plus the standard Condorcet jury theorem analysis. We provide the full derivation.

\paragraph{Part (a): Plurality condition implies convergence.}

Assume $p^* > \max_{y \neq y^*} p_y$, i.e., the correct answer $y^*$ is the unique plurality. Let $\Delta = p^* - \max_{y \neq y^*} p_y > 0$ be the plurality margin.

For majority voting to succeed, we need $y^*$ to appear more frequently than any other answer among the $N$ samples. Let $C_y$ denote the count of answer $y$ among the $N$ samples. Define $Z_i = \mathbf{1}[y_i = y] - \mathbf{1}[y_i = y^*]$, so $C_y - C_{y^*} = \sum_{i=1}^N Z_i$ with $\mathbb{E}[Z_i] = p_y - p^* \leq -\Delta$ and $Z_i \in [-1, 1]$. By Hoeffding's inequality applied to the sum of $N$ variables bounded in $[-1,1]$:
\[
\Pr[C_y > C_{y^*}] = \Pr\!\left[\sum Z_i > 0\right] \leq \Pr\!\left[\sum Z_i - N\mathbb{E}[Z_i] \geq N\Delta\right] \leq \exp\!\left( -\frac{2(N\Delta)^2}{4N} \right) = \exp\!\left( -\frac{N\Delta^2}{2} \right)
\]

By the union bound over all $|Y| - 1$ incorrect answers:
\[
\Pr[\text{majority answer} \neq y^*] \leq (|Y| - 1) \exp\left( -\frac{N\Delta^2}{2} \right)
\]

Thus majority accuracy converges to 1 at rate $\exp(-\Omega(N))$.

\paragraph{Part (b): Correlated errors reduce effective sample size.}

When samples exhibit correlation $\rho$, \Cref{lem:effective_samples} shows that the variance of the sample mean is inflated by factor $(1 + (N-1)\rho)$. This is equivalent to having $N_{\text{eff}} = N / (1 + (N-1)\rho)$ independent samples.

Substituting into the convergence rate from part (a):
\[
\Pr[\text{majority answer} \neq y^*] \leq (|Y| - 1) \exp\left( -\frac{N_{\text{eff}}\Delta^2}{2} \right) = (|Y| - 1) \exp\left( -\frac{N\Delta^2}{2(1 + (N-1)\rho)} \right)
\]

When $\rho = \Omega(1/N)$ or larger, the effective sample size saturates: $N_{\text{eff}} = O(1/\rho)$ regardless of $N$. In this regime, majority voting converges to a fixed (potentially incorrect) answer as $N \to \infty$, because all samples share the same systematic bias.

\paragraph{Part (c): Bottleneck structure determines correlation.}

The logical chain is:
\[
\VC(\calF) \not\subseteq \mathrm{cap}(\calM) \;\xrightarrow{\text{Lemma}~\ref{lem:vc_to_bottleneck}}\; \exists\, V_{\mathrm{sub}} \text{ (bottleneck, Def.~\ref{def:bottleneck})} \;\xrightarrow{\text{Lemma}~\ref{lem:verification_correlation}}\; \rho > 0 \;\xrightarrow{\text{Part (b)}}\; N_{\mathrm{eff}} = O(1/\rho).
\]

\emph{Part (c.i): Shared bottleneck.}
Assume $\VC(\calF) \not\subseteq \mathrm{cap}(\calM)$ and \Cref{asm:dist_hard} holds.
By \Cref{lem:vc_to_bottleneck}, there exists a computational bottleneck $V_{\mathrm{sub}}$ with occurrence probability $q \geq \alpha > 0$ and conditional error rate $r > 1/2$.
If $V_{\mathrm{sub}}$ is a shared bottleneck (every error instance requires $V_{\mathrm{sub}}$), then all $N$ samples from $\calM$ on the same input $x$ face the identical verification failure.
By \Cref{lem:verification_correlation}:
\[
\rho \;\geq\; \frac{q(1-q)(r - p_0)^2}{p_e(1-p_e)} \;>\; 0.
\]
Substituting into part~(b), $N_{\mathrm{eff}} = N / (1 + (N-1)\rho)$.
As $N \to \infty$, $N_{\mathrm{eff}} \to 1/\rho = O(1)$, so additional samples beyond $O(1/\rho)$ provide no further benefit.

\emph{Part (c.ii): Stochastic bottleneck (qualitative prediction).}
This part is an empirically-supported prediction rather than a formally proved result.
When the bottleneck is stochastic, the set of inputs requiring $V_{\mathrm{sub}}$ has measure $q \in (0,1)$ under $D$, and which inputs are hard depends on instance-specific combinatorial structure.
For \emph{between-model} ensembles with models $M_a, M_b$ that have different learned heuristics, the bottleneck instances for $M_a$ and $M_b$ may differ.
Let $q_a$ and $q_b$ be the respective bottleneck probabilities.
The between-model error correlation decomposes as:
\begin{align*}
\rho_{\mathrm{between}} &= \frac{\Pr[E_a \cap E_b] - \Pr[E_a]\Pr[E_b]}{\sqrt{\mathrm{Var}(E_a)\,\mathrm{Var}(E_b)}}.
\end{align*}
When $M_a$ and $M_b$ fail on largely disjoint subsets of instances (because instance-level hardness is stochastic and models use different heuristics), $\Pr[E_a \cap E_b] \approx \Pr[E_a]\Pr[E_b]$, giving $\rho_{\mathrm{between}} \approx 0$.
Heuristically, if the bottleneck instance sets for $M_a$ and $M_b$ are drawn independently with probability $q$, then $\Pr[E_a \cap E_b \mid \text{both bottleneck}] = O(q^2)$ and $\rho_{\mathrm{between}} = O(q^2)$.
A formal bound would require modeling how different architectures and training procedures produce different bottleneck instance sets, which is beyond the scope of this work.
In this regime, between-model majority voting retains its full power: $N_{\mathrm{eff}} \approx N$.

\emph{Part (c.iii): Within-model dominance.}
For within-model sampling from a single model $M$ at temperature $T > 0$, all $N$ samples share $M$'s weights.
If $V_{\mathrm{sub}} \notin \mathrm{cap}(M)$, then on every input $x$ where $V_{\mathrm{sub}}$ is required, every sample from $M$ faces the same computational limitation --- temperature controls token-level randomness but cannot compensate for a missing computational capability.
Therefore, within-model sampling always falls in the shared-bottleneck regime (c.i), regardless of whether the \emph{between-model} regime is shared or stochastic.
The bound $\rho_{\mathrm{within}} > 0$ and $N_{\mathrm{eff}} = O(1/\rho)$ applies.

\paragraph{Connection to the plurality condition.}
The shared bottleneck also interacts with part~(a). When $V_{\mathrm{sub}}$ is shared, the model generates answers via reasoning paths it cannot verify. Since verification failures are systematic, incorrect answers arising from the same bottleneck may appear with probability comparable to $p^*$, potentially violating the plurality condition $p^* > \max_{y \neq y^*} p_y$. Thus self-consistency may fail for two independent reasons:
\begin{enumerate}[label=(\roman*)]
    \item The plurality condition fails (incorrect answer becomes the mode), or
    \item The plurality condition holds but $N_{\mathrm{eff}} = O(1/\rho)$ saturates, bounding the achievable improvement.
\end{enumerate}

This completes the proof of \Cref{thm:selfconsistency}.

\subsection{Proof of Gap Collapse Sharpness}

We show that the gap collapse for HTN planning (\Cref{thm:collapse}) is tight: the verification complexity is exactly $\coNP$-complete for grounded HTN, not higher (e.g., $\PSPACE$-complete), and classical STRIPS planning has strictly easier verification.

\paragraph{Classical STRIPS planning: $\VC = \PTIME$.}

For classical STRIPS planning with grounded operators, a plan $\pi = \langle a_1, \ldots, a_k \rangle$ can be verified in polynomial time:
\begin{enumerate}[label=(\roman*)]
    \item Initialize state $s_0$
    \item For $i = 1, \ldots, k$: check that $a_i$ is applicable in $s_{i-1}$ (preconditions satisfied), then compute $s_i = \gamma(s_{i-1}, a_i)$ by applying effects
    \item Check that the goal holds in $s_k$
\end{enumerate}
Each step takes $O(|\text{state}|)$ time, and there are $k$ steps, so verification is $O(k \cdot |\text{state}|) = \poly(|x|)$.

Thus $\VC(\text{STRIPS}) = \PTIME$.

\paragraph{HTN planning: $\VC = \coNP$-complete for grounded case.}

For HTN planning, a plan is not just a sequence of primitive actions $\pi$, but must be derivable from the initial task network through a sequence of legal task decompositions. Verification requires checking:
\begin{enumerate}[label=(\roman*)]
    \item The action sequence $\pi$ achieves the goal (polynomial-time, as above)
    \item There exists a valid decomposition tree: a tree whose leaves are the actions in $\pi$, whose root is the initial task network, and each internal node represents a legal decomposition of a compound task into subtasks
\end{enumerate}

The second condition is the bottleneck. Given a plan $\pi$ and an HTN problem, the verification question is: ``Does there exist a decomposition tree for $\pi$?''

\citet{behnke2024verification} prove this is $\coNP$-complete by reduction from 3-SAT:
\begin{itemize}
    \item \textbf{Membership in $\coNP$}: The complement problem (``does there \emph{not} exist a valid decomposition tree?'') is in $\NP$. A certificate is a universal quantifier over all possible decomposition trees, checked in polynomial time. Thus the original problem is in $\coNP$.
    \item \textbf{Hardness}: Reduce 3-SAT to HTN plan verification by encoding each clause as a compound task that can be decomposed in multiple ways (one per satisfying assignment). A plan corresponds to a satisfying assignment iff a valid decomposition tree exists.
\end{itemize}

Thus $\VC(\text{HTN}_{\text{grounded}}) = \coNP$.

\paragraph{Tightness: Not $\PSPACE$-complete.}

For \emph{lifted} (non-grounded) HTN planning with variables and quantifiers, verification becomes $\PSPACE$-complete~\citep{behnke2025lifted}. But for grounded HTN (the case relevant to LLM plan generation, where all objects are instantiated), verification is exactly $\coNP$-complete.

This establishes that the hierarchical decomposition structure adds exactly one level of complexity to verification: from $\PTIME$ (STRIPS) to $\coNP$ (grounded HTN). The gap collapse is sharp.

\subsection{Verifying Assumption~\ref{asm:dist_hard} for 3-SAT at the Critical Ratio}
\label{app:assumption_3sat}

We verify that \Cref{asm:dist_hard} (Distributional Verification Hardness) holds concretely for random 3-SAT at the critical clause-to-variable ratio $\alpha \approx 4.27$, connecting directly to our B7 benchmark.

\paragraph{Task setup.}
Consider the 3-SAT decision task $\calF_{\text{3SAT}} = (X, Y, V, D)$ where $X$ is the set of 3-CNF formulas on $n$ variables, $Y = \{\text{SAT}, \text{UNSAT}\} \times \{0,1\}^n$ (a satisfiability verdict plus a candidate assignment), $V(x, y)$ checks: if $y$ claims SAT, verify that the assignment satisfies $x$; if $y$ claims UNSAT, verify that no satisfying assignment exists. The distribution $D$ is the uniform distribution over random 3-SAT instances at clause-to-variable ratio $\alpha$.

\paragraph{Decomposition.}
The verification function decomposes naturally as $V(x,y) = g(V_1(x,y), V_2(x,y))$ where:
\begin{itemize}
    \item $V_1$: \emph{Assignment checking} --- if $y$ claims SAT with assignment $\sigma$, evaluate all clauses of $x$ under $\sigma$. This is in $\PTIME$ (linear in $|\text{formula}|$).
    \item $V_2$: \emph{Unsatisfiability certification} --- if $y$ claims UNSAT, verify that no satisfying assignment exists. This requires deciding membership in $\overline{\text{SAT}}$, which is $\coNP$-complete.
\end{itemize}
The Boolean combination is: $V(x,y) = 1$ iff either ($y$ claims SAT and $V_1 = 1$) or ($y$ claims UNSAT and $V_2 = 1$).

\paragraph{Verifying clause (i): $V_2 \notin \mathrm{cap}(\calM)$.}
For any model class $\calM$ with $\mathrm{cap}(\calM) \subseteq \PTIME$ (including transformers with polynomial-length CoT), computing $V_2$ requires deciding $\coNP$-complete problems. Since $\coNP \not\subseteq \PTIME$ assuming $\PTIME \neq \NP$, we have $V_2 \notin \mathrm{cap}(\calM)$.

\paragraph{Verifying clause (ii): $V_2$ is needed on an $\alpha$-fraction of inputs.}
This is where the phase transition of random 3-SAT is essential. At the critical ratio $\alpha \approx 4.27$:
\begin{itemize}
    \item For $\alpha < 4.27$, almost all random instances are satisfiable (a satisfying assignment exists with high probability as $n \to \infty$).
    \item For $\alpha > 4.27$, almost all random instances are unsatisfiable.
    \item At $\alpha = 4.27$, approximately half of all instances are satisfiable and half are unsatisfiable.
\end{itemize}
This phase transition is rigorously established \citep{mezard2002analytic}.
On the unsatisfiable instances (measure $\approx 1/2$ under $D$), verifying the correct answer (UNSAT) requires $V_2$: the subtask $V_1$ (assignment checking) cannot determine that a formula is unsatisfiable, since there is no assignment to check.
Formally, for an unsatisfiable instance $x$, consider $y_1 = (\text{UNSAT}, \bot)$ (correct) and $y_2 = (\text{SAT}, \sigma)$ for any assignment $\sigma$ (incorrect). We have $V(x, y_1) = 1$ and $V(x, y_2) = 0$, but $V_1(x, y_1) = V_1(x, y_2)$ (both vacuously pass or fail the assignment check depending on convention). Only $V_2$ distinguishes them.

Thus clause (ii) of \Cref{asm:dist_hard} holds with $\alpha \approx 1/2$ --- the unsatisfiability verification subtask is needed on roughly half of all instances at the critical ratio.

\paragraph{Consequence.}
By \Cref{lem:vc_to_bottleneck}, this produces a computational bottleneck $V_{\mathrm{sub}} = V_2$ with occurrence probability $q \approx 1/2$. Since the bottleneck is \emph{stochastic} (whether a particular random formula is unsatisfiable depends on its random clause structure, not on a structural property shared across all instances), this connects directly to \Cref{thm:selfconsistency}(c.ii): different models with different heuristics fail on different unsatisfiable instances, yielding low between-model error correlation. This is consistent with the empirical finding of $\rho = 0.06$ on our B7 (3-SAT) benchmark.

\section{Experimental Details}
\label{app:experiments}

\subsection{Benchmark Construction}

\paragraph{GSM8K.} We use the full test set (1,319 instances) from \citet{cobbe2021gsm8k}. Verification is implemented by evaluating the final numeric answer against the ground truth.

\paragraph{MATH.} We sample 500 instances uniformly from the MATH benchmark \citep{hendrycks2021math} across difficulty levels 1--5. Verification extracts the boxed final answer using regex and compares to ground truth.

\paragraph{3-SAT.} We generate 200 random 3-SAT instances at the critical ratio $m/n = 4.26$ using the standard planted-solutions model. Instances have $n \in [50, 100]$ variables. Verification runs a SAT solver (PySAT) to check satisfiability.

\paragraph{PlanBench-HTN.} We use 100 instances from the hierarchical planning subset of PlanBench \citep{planbench2024}, filtered to tasks requiring at least 3 levels of decomposition. Verification checks: (1) action sequence achieves goal (polynomial-time), (2) valid decomposition tree exists (coNP-complete, verified via exhaustive search with timeout).

\subsection{Model Configuration}

All models use sampling temperature $T = 0.7$ for diversity. We use the following API endpoints:
\begin{itemize}[leftmargin=*]
    \item \textbf{Claude Haiku 4.5}: Anthropic API, model \texttt{claude-haiku-4-5-20251001}
    \item \textbf{GPT-4o-mini}: OpenAI API, model \texttt{gpt-4o-mini-2024-07-18}
    \item \textbf{Llama-3.1-8B}: Together AI API, model \texttt{meta-llama/Llama-3.1-8B-Instruct-Turbo}
\end{itemize}

For each instance, we sample $N \in \{1, 4, 8, 16, 32\}$ independent completions. Majority voting selects the most frequent answer (ties broken randomly).

\subsection{Statistical Analysis}

Accuracy is computed as the fraction of instances where the majority answer matches ground truth. We report 95\% bootstrap confidence intervals (10,000 resamples). Self-consistency lift is the difference Accuracy@32 - Accuracy@1, with significance tested via paired bootstrap test ($p < 0.05$ threshold).

\subsection{Compute Budget}

% TODO: Report actual API costs once experiments complete
Estimated total cost: approximately \$100 across all benchmarks and models (3 models × 4 benchmarks × 300 instances average × 32 samples × \$0.0003 per query average).

\subsection{Additional Results}

% TODO: Add supplementary tables showing:
% - Full breakdown by difficulty level within each benchmark
% - Convergence curves (accuracy vs N) for each model-task pair
% - Error correlation estimates via sample variance analysis
% - Failure mode analysis (qualitative examples of correlated errors on HTN tasks)

% ============================================================

\section*{NeurIPS Paper Checklist}

\begin{enumerate}[leftmargin=*]

\item {\bf Claims.}
\answerYes{Theoretical claims are supported by formal proofs (Theorems 1--3). Empirical validation (Section 5) describes the experimental design and predictions; quantitative results are pending and will be included in a future revision. The paper is a preprint in preparation.}

\item {\bf Limitations.}
\answerYes{Section 7 discusses limitations including restriction to decidable verification functions, worst-case vs average-case gap, the $\TC$ idealization, and the fact that empirical validation is not yet complete.}

\item {\bf Theory Assumptions and Proofs.}
\answerYes{All theorems state assumptions explicitly (standard complexity conjectures). Complete proofs in main text and Appendix A.}

\item {\bf Experimental Result Reproducibility.}
\answerYes{Section 5 specifies models, benchmarks, and sample sizes. Code will be provided in supplementary materials. Quantitative results are pending.}

\item {\bf Open access to data and code.}
\answerYes{All benchmarks are public (GSM8K, MATH, PlanBench). Evaluation code will be provided upon completion of experiments.}

\item {\bf Experimental Setting/Details.}
\answerYes{Section 5 and Appendix B specify the planned experimental protocol. Full results are pending.}

\item {\bf Experiment Statistical Significance.}
\answerYes{The experimental design specifies 95\% bootstrap confidence intervals for all accuracy measurements. Results are pending.}

\item {\bf Experiments Compute Resources.}
\answerYes{Estimated \$100 total API cost. No model training.}

\item {\bf Code Of Ethics.}
\answerYes{}

\item {\bf Broader Impacts.}
\answerYes{Section 6 discusses implications for AI safety and scalable oversight.}

\item {\bf Safeguards.}
\answerNA{Theoretical framework; no high-risk model release.}

\item {\bf Licenses.}
\answerYes{All benchmarks used under their original licenses.}

\item {\bf New Assets.}
\answerNA{No new datasets or models released.}

\item {\bf Crowdsourcing and Human Subjects.}
\answerNA{}

\item {\bf IRB Approvals or Equivalent.}
\answerNA{}

\item {\bf Declaration of LLM usage.}
\answerNA{LLMs are subjects of evaluation, not methodological tools.}

\end{enumerate}

\end{document}
