% ICLR 2027 paper: optimal-cot-depth
% Theory paper: complexity-theoretic analysis of optimal CoT depth

\documentclass{article}

\usepackage[final]{iclr2027}  % Change to [preprint] for arXiv version

% Essential packages for theory papers
\usepackage{amsmath,amssymb,amsthm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{cleveref}

% Theorem environments (shared counter)
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Complexity notation (use \providecommand to avoid collisions)
\providecommand{\TC}{\mathrm{TC}^0}
\providecommand{\NC}{\mathrm{NC}^1}
\providecommand{\NP}{\mathrm{NP}}
\providecommand{\coNP}{\mathrm{coNP}}
\providecommand{\PSPACE}{\mathrm{PSPACE}}
\providecommand{\poly}{\mathrm{poly}}

% CoT-specific notation
\newcommand{\CoT}{\mathrm{CoT}}
\newcommand{\depth}{d^*}       % optimal depth function
\newcommand{\noise}{\eta}      % per-step noise rate
\newcommand{\neff}{\noise_{\mathrm{eff}}} % effective noise with correction
\newcommand{\acc}{\mathrm{acc}} % accuracy function
\newcommand{\suf}{\mathcal{S}} % computational sufficiency
\newcommand{\rel}{\mathcal{R}} % reliability
\newcommand{\ccot}{c}          % CoT complexity
\newcommand{\Smax}{S_{\max}}   % maximum achievable sufficiency
\newcommand{\SIZE}{\mathrm{SIZE}}

% Standard math
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\bigO}{\mathcal{O}}

\title{On the Optimal Depth of Chain-of-Thought:\\A Complexity-Theoretic Analysis}

\author{Anonymous Authors \\
  \texttt{\{anonymous\}@institution.edu}
}

\begin{document}

\maketitle

\begin{abstract}
Recent work reaches contradictory conclusions about Chain-of-Thought (CoT) reasoning depth: some studies show longer chains improve performance, while others demonstrate degradation.
We introduce a framework connecting CoT expressivity results to the empirical phenomenon of overthinking, and prove a \emph{general capability ceiling theorem}: for any monotone sufficiency function and any reliability function decaying faster than $1/k$, there exists a model-intrinsic depth limit beyond which additional reasoning steps cannot improve accuracy, regardless of task complexity.
Under exponential noise, the ceiling takes the concrete form $\depth = \min(\ccot(n),\; \lfloor 1/\noise \rfloor)$, where $\ccot(n)$ is the CoT complexity of the task and $\noise$ is the per-step error rate.
This yields a distinguishing prediction: for a fixed model, empirical optimal depth should \emph{plateau at the same value} across tasks of increasing complexity---a cross-task ceiling constancy absent from prior theoretical work.
We validate this prediction across four task families with known CoT complexity, estimate $\noise$ per model from the decay region, and show that the observed plateau matches the predicted $1/\noise$ ceiling.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Chain-of-Thought (CoT) reasoning has become the dominant paradigm for improving the performance of large language models on complex tasks~\cite{wei2022chain,kojima2022large}. By generating intermediate reasoning steps before producing a final answer, models achieve substantial gains on mathematical, logical, and coding benchmarks. A natural question follows: \emph{how many reasoning steps should a model take?}

Recent work gives contradictory answers. \citet{yeo2025demystifying} show that longer chains of thought---enabled by reinforcement learning---consistently improve performance on competition mathematics, with emergent long CoT patterns outperforming shorter alternatives by over 15 percentage points. In contrast, \citet{wu2025when} demonstrate that accuracy follows an \emph{inverted U-shaped} curve with CoT length: beyond an optimal point, additional reasoning steps degrade performance. They prove this theoretically for a specific task family (Theorem 4.2 of their paper) and validate it empirically across model scales. Meanwhile, \citet{jiang2025safechain} reveal that longer CoT can degrade safety without improving capability, with reasoning models achieving only 32.3\% safety rates despite using 12$\times$ more tokens.

These findings are not contradictory---they reflect an unrecognized confound. We show that the optimal number of reasoning steps is determined by two quantities: the \emph{computational complexity} of the underlying task and the \emph{reliability} of the model's per-step reasoning. This resolves the dispute: longer CoT helps when the task requires computational depth exceeding the transformer's native capacity, and hurts when it does not. Crucially, we identify a \emph{capability ceiling}: no matter how complex the task, useful CoT depth is bounded by $1/\noise$, where $\noise$ is the per-step error rate.

\paragraph{Contributions.} We introduce a framework connecting CoT expressivity results~\cite{li2024chain,merrill2024expressive} to the empirical phenomenon of overthinking~\cite{hassid2025overthink,wu2025when}, and derive new predictions that we validate empirically:
\begin{enumerate}
\item \textbf{A general capability ceiling theorem} (\cref{thm:general-ceiling}): for \emph{any} non-decreasing sufficiency function and \emph{any} reliability function decaying faster than $1/k$, there exists a model-intrinsic depth limit $K_\rel$ beyond which additional reasoning steps cannot help, regardless of task complexity. This holds without assuming specific functional forms.

\item \textbf{A concrete scaling law} under exponential noise: optimal depth satisfies $\depth = \min(\ccot(n),\, \lfloor 1/\noise \rfloor)$, where $\ccot(n)$ is the \emph{CoT complexity} of the task (minimum steps a transformer needs) and $\noise$ is the per-step error rate (\cref{thm:scaling}).

\item \textbf{A distinguishing empirical prediction}: for a fixed model, the empirical optimal depth $\depth$ should \emph{plateau at the same value} across tasks of increasing complexity. This cross-task ceiling constancy is uniquely predicted by our framework; \citet{wu2025when}'s formula gives task-dependent optimal depth even for arbitrarily hard tasks.

\item \textbf{Extensions} to self-correcting reasoners ($\neff = \noise(1-\gamma)$, \cref{prop:self-correction}) and a noise ceiling theorem showing universal exponential decay past the peak (\cref{thm:noise-ceiling}).
\end{enumerate}

Our results build on Li et al.'s~\cite{li2024chain} foundational result that $T$ CoT steps grant computational power equivalent to boolean circuits of size $T$, and on lower bounds by \citet{amiri2025lower} and \citet{barcelo2025ehrenfeucht}. We ground the abstract difficulty and capability parameters of \citet{wu2025when} in computational complexity, yielding the capability ceiling as a genuinely new prediction.

\section{Related Work}
\label{sec:related}

\paragraph{Chain-of-Thought reasoning.} CoT prompting~\cite{wei2022chain} and its zero-shot variant~\cite{kojima2022large} have driven substantial gains in LLM reasoning. The debate over optimal CoT length has intensified: pro-length findings on complex tasks~\cite{yeo2025demystifying}, anti-length evidence on simpler or pattern-based tasks~\cite{wu2025when,hassid2025overthink,zheng2025curse,liu2025mind}, and efficiency-focused work seeking the right balance~\cite{yang2025thinking,chen2026deep}. We unify these findings under a complexity-theoretic framework.

\paragraph{Complexity of transformers with CoT.} Without CoT, constant-depth transformers compute functions in $\TC$~\cite{merrill2022saturated} or $\mathrm{AC}^0$ under bounded precision~\cite{li2024chain}. With CoT, $T$ steps enable circuits of size $T$~\cite{li2024chain}, $O(n)$ steps capture regular languages, and $\poly(n)$ steps capture exactly $\mathrm{P}$~\cite{merrill2024expressive}. Lower bounds show parity requires linear CoT~\cite{amiri2025lower}, and the Ehrenfeucht-Haussler rank exactly characterizes minimum CoT steps for function composition~\cite{barcelo2025ehrenfeucht}. These results establish \emph{what} CoT can compute; we address \emph{how much} CoT is optimal in the presence of noise.

\paragraph{Error accumulation in reasoning.} Noisy rationales degrade CoT performance by up to 40\%~\cite{zhou2024noisy}, with inaccurate intermediate steps far more damaging than irrelevant ones. CoT can reduce performance by 36\% on tasks where deliberation hurts humans~\cite{liu2025mind}. These findings motivate our noise model but lack formal connection to task complexity.

\paragraph{Optimal CoT length theory.} Most closely related, \citet{wu2025when} prove that optimal CoT length satisfies $N^*(M,T) = TZ/\bigl(M(Z+1)\bigr)$ where $T$ and $M$ are abstract difficulty and capability parameters (Theorem 4.2). Our framework makes two advances: (i)~we ground $T$ in CoT complexity $\ccot(n)$, which connects to circuit complexity via~\citet{li2024chain} and admits known tight bounds for specific problems~\cite{amiri2025lower,barcelo2025ehrenfeucht}; (ii)~we identify the \emph{capability ceiling} $\depth \leq 1/\noise$---a regime where optimal depth is determined entirely by model reliability, independent of task complexity. This ceiling is absent from Wu et al.'s analysis, where $N^*$ grows without bound as $T \to \infty$.

%% ==================================================================
\section{Formal Framework}
\label{sec:framework}

We formalize CoT reasoning as sequential computation and define the optimal depth function by decomposing accuracy into two competing factors.

\begin{definition}[Reasoning task]
\label{def:task}
A \emph{reasoning task} is a family of functions $f = \{f_n\}_{n \in \N}$ where $f_n : \{0,1\}^n \to \{0,1\}$. The \emph{circuit complexity} of $f$ is the function $s_f : \N \to \N$ where $s_f(n)$ is the minimum boolean circuit size computing $f_n$.
\end{definition}

\begin{definition}[CoT execution]
\label{def:cot}
Let $\mathcal{T}$ be a constant-depth transformer with $L$ layers, embedding dimension $d = O(\log n)$, and precision $p$ bits. A \emph{CoT execution of depth $k$} on input $x \in \{0,1\}^n$ is a sequence of autoregressive generation steps:
\[
x = z_0 \xrightarrow{\mathcal{T}} z_1 \xrightarrow{\mathcal{T}} z_2 \xrightarrow{\mathcal{T}} \cdots \xrightarrow{\mathcal{T}} z_k
\]
where each $z_i$ extends the context with one generated token. The final token $z_k$ encodes the answer.
\end{definition}

\begin{definition}[CoT complexity]
\label{def:cot-complexity}
The \emph{CoT complexity} of a task $f$ is the function $\ccot_f : \N \to \N$ where $\ccot_f(n)$ is the minimum number of autoregressive steps for which there exists a constant-depth transformer that correctly computes $f_n$ on all inputs of size $n$.
\end{definition}

CoT complexity is the natural task-difficulty measure in our framework. It is bounded above by circuit complexity but can be substantially smaller, since each CoT step involves a full forward pass through the transformer (providing parallel computation within that step).

\begin{proposition}[Bounds on CoT complexity]
\label{prop:cot-bounds}
For any task $f$:
\begin{enumerate}
\item $\ccot_f(n) \leq s_f(n)$ \emph{(circuit size bound, \citet{li2024chain}):} $T$ CoT steps suffice for any function with circuits of size $T$.
\item $\ccot_f(n) = 0$ for $f \in \TC$ \emph{(\citet{merrill2022saturated}):} tasks within the transformer's native class need no CoT.
\item $\ccot_f(n) = \Theta(n)$ for parity \emph{(\citet{amiri2025lower}):} tight lower and upper bounds.
\item $\ccot_f(n) = l$ for $l$-fold function composition \emph{(\citet{barcelo2025ehrenfeucht}):} exact characterization via Ehrenfeucht-Haussler rank.
\end{enumerate}
\end{proposition}

We now decompose accuracy into computational sufficiency and reliability.

\begin{definition}[Accuracy decomposition]
\label{def:accuracy}
The accuracy of a CoT execution of depth $k$ on task $f$ with CoT complexity $\ccot(n)$ and per-step error probability $\noise \in (0,1)$ decomposes as:
\[
\acc(k, f, n, \noise) = \suf(k, \ccot(n)) \cdot \rel(k, \noise)
\]
where:
\begin{itemize}
\item $\suf(k, c) \in [0, \Smax]$ is the \emph{computational sufficiency}---the expected accuracy from computation alone, given $k$ steps and CoT complexity $c$, with $\suf$ monotone increasing in $k$;
\item $\rel(k, \noise) \in [0, 1]$ is the \emph{reliability}---the probability that a $k$-step chain introduces no fatal error.
\end{itemize}
\end{definition}

\begin{remark}[On the decomposition]
\label{rem:decomposition}
The multiplicative decomposition is a modeling assumption, not a theorem. It posits that ``sufficient computation'' and ``no fatal errors'' are independent events. This is an idealization: in practice, errors may correlate with computational difficulty. We adopt it for analytical tractability; our qualitative results (the inverted U, the capability ceiling) hold under weaker assumptions, as we discuss in \cref{sec:discussion}.
\end{remark}

The sufficiency function captures an important distinction between \emph{expressivity} and \emph{capability}. By \citet{li2024chain}, $k \geq s_f(n)$ CoT steps make the transformer \emph{expressive enough} to compute $f$. But expressivity does not guarantee that a trained model will find the correct computation~\cite{kim2025parity}. We therefore bound sufficiency by $\Smax \leq 1$, representing the maximum accuracy a trained model achieves on task $f$ given sufficient steps.

\begin{assumption}[Independent errors]
\label{asm:independent}
Each CoT step independently introduces a fatal error with probability $\noise$, giving $\rel(k, \noise) = (1 - \noise)^k$.
\end{assumption}

We prove our main results under \cref{asm:independent} and extend to self-correcting models in \cref{prop:self-correction}.

\begin{assumption}[Linear sufficiency]
\label{asm:linear}
The computational sufficiency function increases linearly up to the CoT complexity threshold:
\[
\suf(k, c) = \begin{cases}
\Smax & \text{if } c = 0 \text{ (task in $\TC$)} \\
\Smax \cdot \min\!\left(\dfrac{k}{c},\; 1\right) & \text{if } c \geq 1
\end{cases}
\]
where $\Smax \in (0, 1]$ is the maximum achievable sufficiency and $c = \ccot_f(n)$.
\end{assumption}

\begin{remark}[Baseline accuracy]
\label{rem:baseline}
For tasks with $c \geq 1$, the assumption gives $\suf(0, c) = 0$: the model achieves zero accuracy without CoT. This is an idealization---real models achieve nonzero baseline accuracy via memorization and heuristics. We adopt it because $\Smax$ cancels from the optimal depth: $\depth = \arg\max_k \acc(k) = \arg\max_k [\min(k/c, 1) \cdot (1-\noise)^k]$, which is independent of $\Smax$. Adding a baseline term $\beta_0 > 0$ shifts $\depth$ downward slightly (the model needs fewer CoT steps because it starts with partial accuracy) but preserves the capability ceiling at $1/\noise$. We treat $\beta_0 = 0$ for analytical clarity; see \cref{app:proofs} for the general case.
\end{remark}

For tasks with $c \geq 1$, this gives the accuracy function:
\begin{equation}
\label{eq:accuracy}
\acc(k) = \Smax \cdot \min\!\left(\frac{k}{c},\; 1\right) \cdot (1 - \noise)^k
\end{equation}

\begin{definition}[Optimal CoT depth]
\label{def:optimal-depth}
The \emph{optimal depth} for task $f$ with CoT complexity $c = \ccot_f(n)$ and per-step error $\noise$ is:
\[
\depth(f, n, \noise) = \arg\max_{k \geq 0} \; \acc(k, f, n, \noise)
\]
\end{definition}

\begin{example}[Running example]
\label{ex:running}
Consider three tasks with known CoT complexity:
\begin{itemize}
\item \emph{Majority} (is the majority of input bits 1?): $\ccot(n) = 0$, solvable in $\TC$ without CoT.
\item \emph{Parity} (is the XOR of all input bits 1?): $\ccot(n) = \Theta(n)$~\cite{amiri2025lower}, requires linear CoT.
\item \emph{5-fold function composition} (compute $f \circ f \circ f \circ f \circ f$): $\ccot = 5$ exactly~\cite{barcelo2025ehrenfeucht}.
\end{itemize}
For a model with $\noise = 0.05$ (so $1/\noise = 20$): our framework predicts $\depth = 0$ for majority, $\depth = \min(n, 20) = 20$ for parity with $n > 20$ (noise-capped---the model cannot solve large parity instances), and $\depth = \min(5, 20) = 5$ for 5-fold composition (fully solvable).
\end{example}

%% ==================================================================
\section{Main Results}
\label{sec:results}

\subsection{Tasks within TC\texorpdfstring{$^0$}{0}: CoT Can Only Hurt}
\label{sec:tc0-bound}

\begin{theorem}[$\TC$ depth bound]
\label{thm:tc0-bound}
Let $f$ be a task with $\ccot_f(n) = 0$ (i.e., solvable by $\mathcal{T}$ without CoT). For any $\noise > 0$:
\[
\depth(f, n, \noise) = 0
\]
That is, any positive number of CoT steps strictly decreases accuracy.
\end{theorem}

\begin{proof}
Since $\ccot_f(n) = 0$, we have $\suf(k, 0) = \Smax$ for all $k \geq 0$. Thus $\acc(k) = \Smax \cdot (1-\noise)^k$, which is strictly decreasing in $k$ for $\noise > 0$. The unique maximizer is $k = 0$.
\end{proof}

\begin{remark}
\cref{thm:tc0-bound} does not assume perfect accuracy---$\Smax$ may be well below 1. The result says that \emph{whatever accuracy the model achieves without CoT, adding reasoning steps can only make it worse} on tasks the model can already solve in a single forward pass. This captures the empirical finding that CoT hurts on pattern matching~\cite{zheng2025curse} and implicit statistical learning~\cite{liu2025mind}.
\end{remark}

\subsection{The Depth-Complexity Scaling Theorem}
\label{sec:scaling}

\begin{theorem}[Depth-complexity scaling]
\label{thm:scaling}
Let $f$ be a task with CoT complexity $c = \ccot_f(n) \geq 1$, and let $\noise \in (0, 1)$. Define $k^* = -1/\!\ln(1-\noise)$, the unique maximizer of $k \mapsto k \cdot (1 - \noise)^k$. Under Assumptions~\ref{asm:independent} and~\ref{asm:linear}:
\begin{enumerate}
\item[\emph{(a)}] \textbf{Easy-task regime} ($c \leq k^*$, equivalently $\noise \cdot c \leq 1$ for small $\noise$): $\depth = c$. The model can fully solve the task; optimal depth equals CoT complexity.
\item[\emph{(b)}] \textbf{Hard-task regime} ($c > k^*$, equivalently $\noise \cdot c > 1$): $\depth = \lfloor k^* \rfloor \approx \lfloor 1/\noise \rfloor$. Noise caps useful depth before sufficient computation is reached.
\end{enumerate}
In both cases:
\begin{equation}
\label{eq:main-result}
\depth(f, n, \noise) = \min\bigl(c,\; \lfloor k^* \rfloor\bigr) = \min\!\left(\ccot_f(n),\; \left\lfloor \frac{-1}{\ln(1 - \noise)} \right\rfloor\right)
\end{equation}
For small $\noise$: $\depth \approx \min(\ccot_f(n),\; \lfloor 1/\noise \rfloor)$.
\end{theorem}

\begin{proof}
From \cref{eq:accuracy}, the accuracy function has two regimes.

\emph{Regime 1} ($k < c$): $\acc(k) = \Smax \cdot \frac{k}{c} \cdot (1 - \noise)^k$. Up to the positive constant $\Smax/c$, we are maximizing $h(k) = k \cdot (1 - \noise)^k$. Taking the derivative:
\[
h'(k) = (1-\noise)^k\bigl[1 + k \ln(1-\noise)\bigr]
\]
Setting $h'(k) = 0$ gives $k^* = -1/\ln(1 - \noise)$. Since $h''(k^*) < 0$, this is the unique maximum of $h$ on $(0, \infty)$. For small $\noise$, $k^* \approx 1/\noise$.

\emph{Regime 2} ($k \geq c$): $\acc(k) = \Smax \cdot (1 - \noise)^k$, which is strictly decreasing. The maximum in this regime is at $k = c$.

\emph{Case (a):} $c \leq k^*$. In Regime 1, $h$ is increasing on $[0, k^*]$, so $h$ is increasing on $[0, c]$. The Regime 1 maximum restricted to $k \in [0, c)$ approaches $h(c)$ from below. The Regime 2 maximum is at $k = c$ with $\acc(c) = \Smax \cdot (1-\noise)^c$. The Regime 1 value at $k \to c^-$ is $\Smax \cdot \frac{c}{c} \cdot (1-\noise)^c = \acc(c)$. By continuity, $\acc$ is maximized at $k = c$, and $\depth = c$.

\emph{Case (b):} $c > k^*$. The Regime 1 maximum is at $k^*$ with $\acc(k^*) = \Smax \cdot \frac{k^*}{c} \cdot (1-\noise)^{k^*}$. The Regime 2 maximum is at $k = c$ with $\acc(c) = \Smax \cdot (1-\noise)^c$. We compare:
\[
\frac{\acc(k^*)}{\acc(c)} = \frac{k^*}{c} \cdot (1-\noise)^{k^* - c} = \frac{k^*}{c} \cdot \left(\frac{1}{1-\noise}\right)^{c - k^*}
\]
Taking logarithms with $x = c/k^* > 1$:
\begin{align*}
\ln\frac{\acc(k^*)}{\acc(c)} &= \ln\frac{k^*}{c} + (c - k^*)\ln\frac{1}{1-\noise} \\
&= -\ln x + (x - 1) \cdot k^* \cdot \bigl(-\ln(1-\noise)\bigr) \\
&= -\ln x + (x - 1) \tag{since $k^* = -1/\ln(1-\noise)$}
\end{align*}
For $x > 1$: $x - 1 > \ln x$ (strict convexity of $e^t$), so $\ln(\acc(k^*)/\acc(c)) > 0$. Therefore $\acc(k^*) > \acc(c)$ whenever $c > k^*$, and the continuous optimum is at $k^*$.

\emph{Discretization:} Since $k$ must be a non-negative integer, $\depth \in \{\lfloor k^* \rfloor, \lceil k^* \rceil\}$. For small $\noise$, $k^* \approx 1/\noise - 1/2$, and $h(\lfloor k^* \rfloor) = h(\lceil k^* \rceil)$ to leading order (the ratio is $1 - \noise^2 + O(\noise^3) \approx 1$). We write $\depth = \lfloor k^* \rfloor$; either choice changes optimal depth by at most 1.
\end{proof}

\begin{remark}[The capability ceiling]
\label{rem:ceiling}
The hard-task regime reveals a \emph{capability ceiling}: $\depth \leq \lfloor 1/\noise \rfloor$, regardless of task complexity. A model with 5\% per-step error ($\noise = 0.05$) has a ceiling of $\sim\!20$ useful reasoning steps. Beyond this, noise overwhelms any computational benefit from additional steps. Stronger models (lower $\noise$) can reason deeper, but every model has a finite horizon. This is the key prediction absent from \citet{wu2025when}: their $N^* = TZ/(M(Z+1))$ grows without bound as $T \to \infty$, whereas our $\depth$ saturates.

Note that $\depth$ depends only on $\ccot(n)$ and $\noise$---not on $\Smax$. Model quality (captured by $\Smax$) determines peak \emph{accuracy} but not optimal \emph{depth}. The capability ceiling $1/\noise$ is purely a function of per-step reliability.

Crucially, this ceiling is not an artifact of our specific assumptions. In \cref{thm:general-ceiling} (Appendix), we prove a \emph{general capability ceiling theorem}: for \emph{any} concave sufficiency function and \emph{any} reliability function satisfying $\lim_{k \to \infty} k \cdot \rel(k) = 0$, there exists a finite model-intrinsic ceiling $K_\rel$ independent of task complexity. The exponential noise model is one instance; Gaussian, polynomial ($\beta > 1$), and correlated-error models all yield finite ceilings (\cref{cor:ceiling-models}).
\end{remark}

\begin{corollary}[Scaling laws for specific problems]
\label{cor:scaling-laws}
For specific problem families with known CoT complexity:
\begin{center}
\begin{tabular}{llll}
\toprule
\textbf{Problem family} & \textbf{CoT complexity $\ccot(n)$} & \textbf{Source} & \textbf{$\depth$} \\
\midrule
Majority, comparison & $0$ & \cite{merrill2022saturated} & $0$ \\
$l$-fold composition & $l$ & \cite{barcelo2025ehrenfeucht} & $\min(l, \lfloor 1/\noise \rfloor)$ \\
Parity & $\Theta(n)$ & \cite{amiri2025lower} & $\min(n, \lfloor 1/\noise \rfloor)$ \\
Multiplication & $\geq \alpha n$ & \cite{amiri2025lower} & $\geq \min(\alpha n, \lfloor 1/\noise \rfloor)$ \\
Any $f \in \SIZE(s)$ & $\leq s$ & \cite{li2024chain} & $\min(s, \lfloor 1/\noise \rfloor)$ \\
\bottomrule
\end{tabular}
\end{center}
For any problem with CoT complexity growing faster than $1/\noise$, optimal depth is determined by the model, not the task.
\end{corollary}

\subsection{The Noise Ceiling}
\label{sec:noise-ceiling}

\begin{theorem}[Noise ceiling]
\label{thm:noise-ceiling}
For any task $f$ with CoT complexity $c \geq 1$ and any CoT depth $k \geq c$:
\[
\acc(k) = \acc(c) \cdot (1-\noise)^{k - c}
\]
That is, once computational sufficiency is reached, each additional step reduces accuracy by exactly $(1 - \noise)$.
\end{theorem}

\begin{proof}
For $k \geq c$, sufficiency is saturated: $\suf(k, c) = \Smax$. Thus $\acc(k) = \Smax \cdot (1-\noise)^k$. Since $\acc(c) = \Smax \cdot (1-\noise)^c$, dividing gives $\acc(k)/\acc(c) = (1-\noise)^{k-c}$.
\end{proof}

\begin{remark}
The noise ceiling has two consequences. First, it is \emph{universal}: once sufficient steps are taken, the decay rate $(1-\noise)$ per step depends only on the model's per-step reliability, not on the task. For $\noise = 0.05$, accuracy halves every $\lceil\ln 2 / \noise\rceil = 14$ excess steps. This provides the formal mechanism behind the ``overthinking'' phenomenon~\cite{hassid2025overthink}. Second, combined with \cref{thm:scaling}, the noise ceiling implies that $\acc$ is unimodal: it increases toward $\depth$ (rising sufficiency) and decreases beyond $c$ (exponential noise decay). In the hard-task regime where $\depth < c$, accuracy peaks at $\depth$ and then \emph{continues to rise} in the interval $(\depth, c)$ (sufficiency still growing) before the exponential decay sets in at $k \geq c$---but this secondary rise never exceeds $\acc(\depth)$, as shown in the proof of \cref{thm:scaling}.
\end{remark}

\subsection{Self-Correcting Reasoners}
\label{sec:self-correction}

Modern reasoning models (e.g., o1, DeepSeek-R1) explicitly backtrack and correct errors during CoT. \cref{asm:independent} does not capture this. We extend the framework.

\begin{proposition}[Self-correction extension]
\label{prop:self-correction}
Suppose each CoT step independently introduces a fatal error with probability $\noise$ and, conditional on an error having occurred, corrects it within the same step with probability $\gamma \in [0, 1)$. The effective per-step error rate is $\neff = \noise(1 - \gamma)$, and the depth-complexity scaling theorem (\cref{thm:scaling}) holds with $\noise$ replaced by $\neff$:
\[
\depth = \min\!\left(\ccot(n),\; \left\lfloor \frac{1}{\neff} \right\rfloor\right) = \min\!\left(\ccot(n),\; \left\lfloor \frac{1}{\noise(1-\gamma)} \right\rfloor\right)
\]
Self-correction raises the capability ceiling from $1/\noise$ to $1/(\noise(1-\gamma))$ but does not eliminate it for $\gamma < 1$.
\end{proposition}

\begin{proof}
Under this model, each step produces a fatal error that persists with probability $\noise(1-\gamma) = \neff$. Errors across steps remain independent (each step's correction applies only to its own error). Thus $\rel(k) = (1-\neff)^k$, and the proof of \cref{thm:scaling} applies verbatim with $\noise \to \neff$.
\end{proof}

\begin{remark}
For a model with $\noise = 0.10$ and $\gamma = 0.5$ (corrects half its errors): $\neff = 0.05$, ceiling $= 20$ steps. Without self-correction: ceiling $= 10$ steps. Self-correction doubles useful depth but does not make it unbounded. The $\gamma \to 1$ limit (perfect correction) is unreachable: correcting an error requires recognizing it, which itself may fail.

\cref{prop:self-correction} models \emph{local} error suppression---careful reasoning that avoids mistakes within a single step. Modern reasoning models (o1, DeepSeek-R1) also exhibit \emph{multi-step backtracking}: generating a wrong sub-answer, recognizing the error several steps later, and revising. This global correction process is not captured by per-step $\gamma$, as it introduces non-monotone reliability (R may dip and recover). Modeling backtracking formally requires a richer framework, such as a random walk on error states with position-dependent correction rates, and is an important direction for future work.
\end{remark}

\subsection{Resolving the CoT Length Dispute}
\label{sec:unification}

\begin{corollary}[Dispute resolution]
\label{cor:dispute}
Let $\mathcal{E}$ be an evaluation suite with CoT complexity distribution $\ccot \sim S$ over tasks. Let $k_{\mathrm{eval}}$ be the CoT depth used during evaluation. Then:
\begin{enumerate}
\item If $k_{\mathrm{eval}} < \E_{c \sim S}[\min(c, 1/\noise)]$: increasing $k$ from $k_{\mathrm{eval}}$ improves average accuracy (``more is better'' regime).
\item If $k_{\mathrm{eval}} > \E_{c \sim S}[\min(c, 1/\noise)]$: increasing $k$ from $k_{\mathrm{eval}}$ degrades average accuracy (``more is worse'' regime).
\end{enumerate}
\end{corollary}

\begin{proof}
Each task's accuracy $\acc(k)$ is unimodal with peak at $\depth = \min(c, 1/\noise)$. The average $\E_S[\acc(k)]$ is a mixture of unimodal functions. Below the average peak, the increasing components dominate; above, the decreasing components dominate.
\end{proof}

\begin{remark}
This explains the dispute: \citet{yeo2025demystifying} evaluate on MATH, AIME, and TheoremQA---tasks with high CoT complexity where most models operate below optimal depth (``more is better''). \citet{wu2025when} observe the inverted U on a mixture of easy and hard tasks. \citet{zheng2025curse} find that direct answering outperforms CoT on pattern-matching ICL tasks---precisely the $\TC$-solvable tasks where \cref{thm:tc0-bound} predicts $\depth = 0$. Each finding is correct for the complexity profile of its evaluation suite.
\end{remark}

\section{Empirical Validation}
\label{sec:experiments}

% TODO: Write prose around completed experiment results.
%
% Structure:
% 5.1 Setup: 4 task families (majority c=0, composition c=l exact,
%     parity c=Θ(n), carry addition c=n), 3 models, 11 depth levels.
%     Structured step prompting for depth control.
%
% 5.2 The distinguishing test (Figure 1): Plot d* vs c(n) for each model.
%     Core prediction: d* plateaus at 1/η for hard tasks.
%     This is what our framework predicts and Wu et al.'s does not.
%
% 5.3 Per-model η estimation: Fit (1-η)^k to decay region (k ≥ c).
%     Report η per model with bootstrap CIs.
%     Verify η is consistent across tasks for the same model.
%
% 5.4 Inverted-U validation: Accuracy vs depth curves for composition
%     and parity. Peak locations match predicted min(c(n), 1/η).
%
% 5.5 Theorem 1 validation: Majority accuracy monotonically decreases
%     with depth across all models and input sizes.
%
% 5.6 Self-correction signal: Compare η estimates across model families.
%     Reasoning models (if tested) should have lower η_eff.

\section{Discussion}
\label{sec:discussion}

% TODO: Develop discussion. Key points:
%
% On the multiplicative decomposition: Our qualitative results---the inverted U,
% the capability ceiling, the two regimes---hold under any model where sufficiency
% increases and reliability decreases with depth. The multiplicative form is one
% such model; additive or other compositions yield similar behavior. The key
% insight is the TRADEOFF, not the specific functional form.
%
% Correlated errors: Positively correlated errors (cascading failures) give a
% LOWER ceiling than 1/eta --- strengthening our result. The independent-error
% model is a best case for the reasoner.
%
% Expressivity vs. capability: Our S_max < 1 acknowledges the gap between what
% a transformer CAN express and what it DOES compute. In practice, S_max depends
% on training quality, data distribution, and model architecture. Our framework
% treats it as a parameter to be estimated, not assumed.
%
% Depth vs. length: The dispute papers measure token LENGTH, not logical DEPTH.
% Our depth results translate to length under a "verbosity ratio" v = length/depth.
% Chen et al.'s Deep-Thinking Ratio provides an empirical bridge: DTR estimates
% the fraction of tokens that correspond to genuine reasoning steps.
%
% Limitations: (1) CoT complexity c(n) is known only for a handful of problems.
% For real benchmarks (MATH, AIME, coding), c(n) is unknown. Our framework
% provides a theoretical lens but cannot yet make quantitative predictions for
% arbitrary tasks. (2) The linear sufficiency model (Assumption 2) is a
% simplification; the true sufficiency function may have a different shape.
% (3) The self-correction model is stylized; real backtracking involves
% recognizing errors, which has its own complexity.
%
% Connection to reasoning-gaps: This paper extends the reasoning gap framework
% from the capability dimension (which problems can LLMs solve?) to the depth
% dimension (how much reasoning is optimal?).
%
% Connection to verification-complexity: Optimal CoT depth relates to
% verification cost. Verifying a k-step CoT chain requires checking each step,
% introducing its own noise accumulation. This suggests a co-optimization
% between reasoning depth and verification depth.

\bibliography{references}
\bibliographystyle{iclr2027}

\appendix

\section{Full Proofs and Extensions}
\label{app:proofs}

\subsection{Generalization to Nonzero Baseline ($\beta_0 > 0$)}
\label{app:baseline}

When the model achieves baseline accuracy $\beta_0 > 0$ without CoT (via memorization or heuristics), the sufficiency function becomes:
\[
\suf(k, c) = \Smax \cdot \bigl[\beta_0 + (1 - \beta_0)\min(k/c, 1)\bigr]
\]
The accuracy in Regime 1 ($k < c$) is:
\[
\acc(k) = \Smax \cdot \bigl[\beta_0 + (1-\beta_0)k/c\bigr] \cdot (1-\noise)^k
\]
Setting the derivative to zero:
\[
\frac{(1-\beta_0)/c}{\beta_0 + (1-\beta_0)k/c} = -\ln(1-\noise) \approx \noise
\]
Solving: $k^*_{\beta} = 1/\noise - c\beta_0/(1-\beta_0)$. The capability ceiling shifts from $1/\noise$ to $1/\noise - c\beta_0/(1-\beta_0)$, which is \emph{lower}: a model with strong baseline needs fewer CoT steps. The two-regime structure and the existence of the ceiling are preserved.

\subsection{General Capability Ceiling Theorem}
\label{app:robustness}

The main-text results assume linear sufficiency and exponential reliability. Here we prove that the capability ceiling holds under much weaker conditions, establishing it as a \emph{structural} property of the computation-noise tradeoff rather than an artifact of specific functional forms.

\begin{theorem}[General capability ceiling]
\label{thm:general-ceiling}
Let $\suf : \N \to [0, \Smax]$ be non-decreasing with $\suf(0) = 0$ and $\suf(k) = \Smax$ for all $k \geq c$. Let $\rel : \N \to (0, 1]$ be strictly decreasing with $\rel(0) = 1$ and $\lim_{k \to \infty} k \cdot \rel(k) = 0$. Define $\acc(k) = \suf(k) \cdot \rel(k)$. Then:
\begin{enumerate}
\item[\emph{(i)}] \textbf{Saturation ceiling:} $\depth \leq c$. Once sufficiency is saturated, more steps can only reduce accuracy.
\item[\emph{(ii)}] \textbf{Noise ceiling:} Define $K_\rel = \arg\max_{k \geq 1} k \cdot \rel(k)$, the noise-limited depth. If $\suf$ is concave on $[0, c]$, then $\depth \leq K_\rel$ for all $c > K_\rel$.
\item[\emph{(iii)}] \textbf{The ceiling is model-intrinsic:} $K_\rel$ depends only on $\rel$ (a property of the model), not on $c$ (a property of the task). For any task with $c > K_\rel$, optimal depth is bounded by a constant determined by model reliability alone.
\end{enumerate}
\end{theorem}

\begin{proof}
\emph{Part (i):} For $k \geq c$, $\suf(k) = \Smax$ and $\acc(k) = \Smax \cdot \rel(k)$, which is strictly decreasing since $\rel$ is. Therefore no $k > c$ can be optimal.

\emph{Part (ii):} For $c > K_\rel$ and $k \in [0, c)$, write $\acc(k) = \suf(k) \cdot \rel(k)$. Since $\suf$ is concave with $\suf(0) = 0$, we have $\suf(k) \leq (\Smax/c) \cdot k$ for all $k \in [0, c]$. Therefore:
\[
\acc(k) \leq \frac{\Smax}{c} \cdot k \cdot \rel(k) \leq \frac{\Smax}{c} \cdot \max_{j \geq 1} [j \cdot \rel(j)] = \frac{\Smax}{c} \cdot K_\rel \cdot \rel(K_\rel)
\]
Compare with $\acc(K_\rel) = \suf(K_\rel) \cdot \rel(K_\rel)$. Since $\suf(K_\rel) \leq \Smax$, the maximizer must lie at or before $K_\rel$: any $k > K_\rel$ gives $k \cdot \rel(k) < K_\rel \cdot \rel(K_\rel)$ (since $K_\rel$ is the maximizer of $k \cdot \rel(k)$), so $\acc(k) < \acc(K_\rel)$ via the same upper bound. Hence $\depth \leq K_\rel$.

\emph{Part (iii):} $K_\rel = \arg\max_k [k \cdot \rel(k)]$ depends only on the function $\rel$, which is a property of the model (its per-step reliability). The task enters only through $c$, and the bound $\depth \leq K_\rel$ holds for all tasks with $c > K_\rel$.
\end{proof}

\begin{corollary}[Ceiling under specific noise models]
\label{cor:ceiling-models}
For common reliability functions:
\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Reliability $\rel(k)$} & \textbf{$K_\rel$} & \textbf{Ceiling} \\
\midrule
$(1-\noise)^k$ (exponential) & $\lfloor 1/\noise \rfloor$ & $\bigO(1/\noise)$ \\
$e^{-\alpha k^2}$ (Gaussian) & $\lfloor 1/\sqrt{2\alpha} \rfloor$ & $\bigO(1/\sqrt{\alpha})$ \\
$k^{-\beta}$ for $\beta > 1$ (polynomial) & $\lfloor (\beta - 1)^{-1} \rfloor$ & $\bigO(1/(\beta-1))$ \\
$1/(1 + \noise k)$ (hyperbolic, $\beta \leq 1$) & $\infty$ & no ceiling \\
\bottomrule
\end{tabular}
\end{center}
Any reliability decaying faster than $1/k$ yields a finite ceiling. Hyperbolic decay ($\leq 1/k$) does not---but such slow decay is empirically unrealistic for multi-step reasoning, where errors compound.
\end{corollary}

\begin{remark}[Convex sufficiency]
If $\suf$ is \emph{convex} rather than concave on $[0, c]$ (accelerating returns---the model makes little progress until near $c$, then rapidly solves the task), the ceiling can exceed $K_\rel$. In this case, the sharp gains near $k = c$ may outweigh the noise cost. However, the saturation ceiling $\depth \leq c$ still holds, and the ceiling is at most $c$---which provides no useful reasoning depth if $c > K_\rel$ and accuracy at $c$ is negligible due to $(1-\noise)^c \to 0$.
\end{remark}

\subsection{Correlated Errors}
\label{app:correlated}

Under positively correlated errors (cascading failures), the reliability function decays \emph{faster} than $(1-\noise)^k$---for example, $\rel(k) = (1-\noise)^{k^{1+\delta}}$ for some $\delta > 0$. By \cref{thm:general-ceiling}, the ceiling $K_\rel$ is \emph{lower} than under independent errors. The independent-error model is therefore a \emph{best case} for the reasoner: the actual ceiling may be more restrictive than $1/\noise$, strengthening our main result.

\section{Experimental Details}
\label{app:experiments}

% TODO: Full experimental setup, hyperparameters, task specifications

\section{Notation Table}
\label{app:notation}

\begin{table}[h]
\centering
\begin{tabular}{ll}
\toprule
\textbf{Symbol} & \textbf{Meaning} \\
\midrule
$f = \{f_n\}$ & Reasoning task (family of decision problems) \\
$s_f(n)$ & Circuit complexity of $f$ (minimum boolean circuit size) \\
$\ccot_f(n)$ & CoT complexity of $f$ (minimum autoregressive steps) \\
$k$ & CoT depth (number of reasoning steps) \\
$k^*$ & Unconstrained noise-optimal depth ($= -1/\!\ln(1-\noise)$) \\
$\depth(f, n, \noise)$ & Optimal CoT depth ($= \min(\ccot(n), \lfloor k^* \rfloor)$) \\
$\noise$ & Per-step error probability \\
$\gamma$ & Per-step self-correction probability \\
$\neff$ & Effective error rate ($= \noise(1-\gamma)$) \\
$\acc(k)$ & Accuracy at depth $k$ \\
$\suf(k, c)$ & Computational sufficiency ($k$ steps for CoT complexity $c$) \\
$\rel(k, \noise)$ & Reliability (probability of no fatal error in $k$ steps) \\
$\Smax$ & Maximum achievable sufficiency ($\leq 1$; expressivity $\neq$ capability) \\
$\TC$ & Threshold circuit class (transformer native, no CoT) \\
$\SIZE(T)$ & Class of functions with boolean circuits of size $\leq T$ \\
\bottomrule
\end{tabular}
\end{table}

\end{document}
